### 2014-02-19 paper/paper.tex
The initial, empty paper

\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}

\title{Lock-free skiplist-based priority queues}
\author{Jakob Gruber, 0203440}

\begin{document}

\maketitle

\begin{abstract}
blablablu blublubla
\end{abstract}

\section{Introduction}
\section{Related Work}
\section{Conclusion}

\end{document}

### 2014-02-19 paper/paper.tex
Packages, bibliography, acronyms, etc.
Abstract and introduction formula

@@
 
+\usepackage{acronym}
+\usepackage[backend=bibtex]{biblatex}
+\usepackage{comment}
+\usepackage[pdfborder={0 0 0}]{hyperref}
 \usepackage[utf8]{inputenc}
+
+\input{../common/acronyms}
+\bibliography{../common/bibliography.bib}
 
@@
 \maketitle
+
+\begin{comment}
+Abstract formula
+----------------
+
+1) big picture problem or topic widely debated in your field.
+2) gap in the literature on this topic.
+3) your project filling the gap.
+4) the specific material that you examine in the paper.
+5) your original argument.
+6) a strong concluding sentence.
+
+Introduction formula
+--------------------
+
+A good paper introduction is fairly formulaic. If you follow a simple set of
+rules, you can write a very good introduction. The following outline can be
+varied. For example, you can use two paragraphs instead of one, or you can
+place more emphasis on one aspect of the intro than another. But in all cases,
+all of the points below need to be covered in an introduction, and in most
+papers, you don't need to cover anything more in an introduction.
+
+Paragraph 1: Motivation. At a high level, what is the problem area you are
+working in and why is it important? It is important to set the larger context
+here. Why is the problem of interest and importance to the larger community?
+
+Paragraph 2: What is the specific problem considered in this paper? This
+paragraph narrows down the topic area of the paper. In the first paragraph you
+have established general context and importance. Here you establish specific
+context and background.
+
+Paragraph 3: "In this paper, we show that ...". This is the key paragraph in
+the intro - you summarize, in one paragraph, what are the main contributions of
+your paper given the context you have established in paragraphs 1 and 2. What
+is the general approach taken? Why are the specific results significant? This
+paragraph must be really really good. If you can't "sell" your work at a high
+level in a paragraph in the intro, then you are in trouble. As a reader or
+reviewer, this is the paragraph that I always look for, and read very
+carefully.
+
+You should think about how to structure this one or two paragraph summary of
+what your paper is all about. If there are two or three main results, then you
+might consider itemizing them with bullets or in test (e.g., "First, ..."). If
+the results fall broadly into two categories, you can bring out that
+distinction here. For example, "Our results are both theoretical and applied in
+nature. (two sentences follow, one each on theory and application)"
+
+Paragraph 4: At a high level what are the differences in what you are doing,
+and what others have done? Keep this at a high level, you can refer to a future
+section where specific details and differences will be given. But it is
+important for the reader to know at a high level, what is new about this work
+compared to other work in the area.
+
+Paragraph 5: "The remainder of this paper is structured as follows..." Give the
+reader a roadmap for the rest of the paper. Avoid redundant phrasing, "In
+Section 2, In section 3, ... In Section 4, ... " etc.
+
+A few general tips:
+
+Don't spend a lot of time into the introduction telling the reader about what
+you don't do in the paper. Be clear about what you do do, but don't dwell here
+on what you don't do.  Does each paragraph have a theme sentence that sets the
+stage for the entire paragraph? Are the sentences and topics in the paragraph
+all related to each other?  Do all of your tenses match up in a paragraph?
+\end{comment}
 
@@
 
+% TODO: Remove me.
+\nocite{*}
+
+\printbibliography
+
 \end{document}

### 2014-02-19 paper/paper.tex
More information in paper title
Further notes

@@
 
-\title{Lock-free skiplist-based priority queues}
+\title{Lock-free Skiplist-based Priority Queues \\
+       Student Paper for Seminar in Algorithms 2013W \\
+       Technical University of Vienna}
 \author{Jakob Gruber, 0203440}
@@
 \section{Introduction}
+
+\begin{comment}
+Sections / rough structure:
+* Basic concepts and definitions. Linearizability, sequ./quiescent consistency,
+  lock-free, wait-free, disjoint-access parallelism (one of the papers has good
+  summaries of these).
+  Maybe atomic primitives such as CAS (but probably not).
+* Priority queue definitions, semantics, usages. Mention inherent
+  non-scalability through DeleteMin().
+* Skiplist, heap definitions, semantics, usages.
+* Follow development of current state of the art from Hunt Heap ~> Shavit ~>
+  Tsigas ~> Linden.
+* Maybe benchmarks on mars.
+\end{comment}
+
 \section{Related Work}
+
+\begin{comment}
+  * hunt: most efficient old-school algorithm
+  * lots of other older references in [4]
+  * israeli, rappoport: wait-free, non-available atomic primitive
+  * lotan, shavit [4]: skiplist, lock-based. first to propose usage of skip-lists [11]
+  * sundell, tsigas [3]: skiplist, lock-free, linearizable, unique priorities
+  * herlihy, shavit in art of multiprocessor programming: based on [4], lock-free.
+  * linden, jonsson [11]: skiplist, lock-free, linearizable, reduced mem contention.
+    first lock-free PQ algorithm [11, 12]
+  * specialized versions such as
+    * bounded priorities: [7] and others
+    * probabilistic extractMin, extractMany: [9]
+\end{comment}
+
 \section{Conclusion}
+
+\begin{comment}
+http://leo.stcloudstate.edu/acadwrite/conclude.html
+http://writingcenter.unc.edu/handouts/conclusions/
+http://www.wikihow.com/Write-a-Conclusion-for-a-Research-Paper
+\end{comment}
 

### 2014-02-19 paper/paper.tex
Introduction

@@
 \usepackage[backend=bibtex]{biblatex}
+\usepackage[usenames,dvipsnames]{color}
 \usepackage{comment}
@@
 \usepackage[utf8]{inputenc}
+\usepackage{listings}
 
@@
 \bibliography{../common/bibliography.bib}
+
+\definecolor{Gray}{gray}{0.5}
+\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
+
+\lstset{
+    language=C++,
+    basicstyle=\ttfamily,
+    keywordstyle=\color{OliveGreen},
+    commentstyle=\color{Gray},
+    captionpos=b,
+    breaklines=true,
+    breakatwhitespace=false,
+    showspaces=false,
+    showtabs=false,
+    numbers=left,
+}
 
@@
 
+In the past decade, advancements in computer performance have been made mostly
+through increasing the number of processors instead of higher clock speeds.
+This development necessitates new approaches to data structures and algorithms
+that take advantage of concurrent execution on multiple threads and processors.
+
+This paper focuses on the priority queue data structure, consisting of two operations
+traditionally called \lstinline|Insert| and \lstinline|DeleteMin|. \lstinline|Insert|
+places an item together with its priority into the queue, while \lstinline|DeleteMin|
+removes and returns the highest priority item. Both operations are expected to have
+a complexity of at most $O(\log n)$. Priority queues are used in a large variety
+of situations such as shortest path algorithms and scheduling.
+
+Concurrent priority queues have been the subject of research since the 1980s.
+While early efforts have focused mostly on parallelizing Heap structures
+\cite{hunt1996efficient}, % TODO: Many more, see \cite{shavit2000skiplist}
+more recently priority queues based on \citeauthor{pugh1990skip}'s SkipLists
+\cite{pugh1990skip} seem to show more potential \cite{shavit2000skiplist,
+sundell2003fast,herlihy2012art,linden2013skiplist}.
+
+In the following, we examine the evolution of concurrent priority queues from
+an early prominent Heap-based algorithm through initial SkipList-based designs
+to current state-
+
 \begin{comment}

### 2014-02-20 paper/paper.tex
A sample abstract text

@@
 \begin{abstract}
-blablablu blublubla
+On the 24th of February, 1815, the watch-tower of Notre-Dame de la Garde
+signalled the arrival of the three-master Pharaon, from Smyrna, Trieste, 
+and Naples.
+
+The usual crowd of curious spectators immediately filled the quay of Fort 
+Saint-Jean, for at Marseilles the arrival of a ship is always a great event, 
+especially when that ship, as was the case with the Pharaon, has been built, 
+rigged, and laden in the dockyard of old Phocaea and belongs to a shipowner 
+of their own town.
+
+Meanwhile the vessel drew on, and was approaching the harbour under topsails, 
+jib, and foresail, but so slowly and with such an air of melancholy that the 
+spectators, always ready to sense misfortune, began to ask one another what 
+ill-luck had overtaken those on board. However, those experienced in navigation 
+soon saw that if there had been any ill-luck, the ship had not been the 
+sufferer, for she advanced in perfect condition and under skilful handling; 
+the anchor was ready to be dropped, the bowsprit shrouds loose. 
+Beside the pilot, who was steering the Pharaon through the narrow entrance to 
+the port, there stood a young man, quick of gesture and keen of eye, who 
+watched every movement of the ship while repeating each of the pilot's orders.
 \end{abstract}

### 2014-02-20 paper/paper.tex
Notes, concepts, definitions

@@
 
+% Move to many cores, why do we need concurrent algorithms/data structures?
+
 In the past decade, advancements in computer performance have been made mostly
-through increasing the number of processors instead of higher clock speeds.
+through an increasing number of processors instead of higher clock speeds.
 This development necessitates new approaches to data structures and algorithms
@@
 
+% The priority queue - semantics and usage summary.
+
 This paper focuses on the priority queue data structure, consisting of two operations
 traditionally called \lstinline|Insert| and \lstinline|DeleteMin|. \lstinline|Insert|
-places an item together with its priority into the queue, while \lstinline|DeleteMin|
-removes and returns the highest priority item. Both operations are expected to have
+places an item into the queue together with its priority, while \lstinline|DeleteMin|
+removes and returns the highest priority item. Both of these operations are expected to have
 a complexity of at most $O(\log n)$. Priority queues are used in a large variety
-of situations such as shortest path algorithms and scheduling.
+of situations such as shortest path algorithms and scheduling. % See liu2012lock for more.
+
+% Concurrent priority queues - heaps, trees, skiplists.
 
@@
 
-In the following, we examine the evolution of concurrent priority queues from
-an early prominent Heap-based algorithm through initial SkipList-based designs
-to current state-
+% Our plan for the paper - somehow this kind of duplicates the last paragraph though.
+
+In the following, we examine the evolution of concurrent priority queues. We begin
+with first steps which used fine-grained locking to avoid the bottleneck of a single
+global lock. Initial SkipList-based structures offered better disjoint-access
+parallelism, and finally current state of the art lock-free priority queues minimize
+contention in \lstinline|DeleteMin|.
+
+% TODO: In section x, we ... Again, this partially duplicates the last paragraph.
+
+\section{Concepts and Definitions}
+
+% Blocking, lock-free and wait-free data structures.
+
+Concurrent data structures are intended to be accessed simultaneously by several processes
+at once. \emph{Lock-based} structures ensure that only a limited number of processes may enter
+a critical section at once. \emph{Lock-free} data structures eschew the use of locks, and guarantee
+that at least a single process makes progress at all times. Since lock-free structures are
+non-blocking, they are not susceptible to priority inversion, deadlock, and livelock.
+\emph{Wait-freedom} further guarantees that every process finishes each operation in a bounded number of steps. 
+In practise, wait-freedom often introduces an unacceptable overhead; lock-freedom
+however has proven to be both efficient and to scale well to large numbers of processes.
+
+% Linearizability, sequential and quiescent consistency.
+
+There are several different criteria which allows reasoning about the correctness of concurrent
+data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
+effect at a single instant in time at so-called linearization points. Linearizability is composable ---
+any data structure composed of linearizable objects is also linearizable.
+\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
+executed in parallel to be equivalent to the result of some sequential ordering of the same
+operations.
+\emph{Quiescently consistent} \cite{shavit1996diffracting}
+guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
+after a period of quiescence, i.e. an interval without active operations has passed.
+
+% Disjoint-access parallelism.
+
+\section{Fine-grained Locking Heaps}
+\section{SkipList-based Priority Queues}
 

### 2014-02-20 paper/paper.tex
Further work on concepts and definitions

@@
 
+An area in memory accessed frequently by a large number of processes is said to be \emph{contended}.
+Contention is an limiting factor regarding scalability: concurrent reads and writes to the same 
+location must be serialized by the cache coherence protocol, and only a single concurrent \ac{CAS} % TODO: Short section on primitives.
+can succeed while all others must retry. \emph{Disjoint-access parallelism} is the concept of
+spreading such accesses in order to reduce contention as much as possible.
+
+% Priority queues.
+
+Priority queues are an abstract data structure allowing insertion an items with a given priority
+and removal of the highest-priority item in logarithmic time. Search trees and Heaps (which are 
+flattened representations complete trees such that each node's value is at least as large as
+those of both children) are concrete data structures which are usually used to implement sequential
+priority queues. However, both require fairly elaborate reorganization after \lstinline|DeleteMin|
+and/or \lstinline|Insert| operations, which are especially challenging to achieve in a concurrent
+environment.
+
+% SkipLists.
+
+SkipLists \cite{pugh1990skip} have become increasingly popular in concurrent data structures because 
+they are both simple and exhibit excellent disjoint-access parallelism. In contrast to
+both search trees and Heaps, reorganization is not necessary since SkipLists rely on
+randomization for an expected $O(\log n)$ complexity of \lstinline|Insert|, \lstinline|Search|
+and \lstinline|Delete| operations.
+A SkipList may be visualized as a set of linked lists with corresponding levels. The linked
+list at level 0 is the sorted sequence of all objects in the SkipList,
+and higher levels provide ``shortcuts''
+into the SkipList such that a list of level $i + 1$ contains a subset of the objects in level
+$i$. A SkipList node $n$ is said to be of level $i$ if it is in all lists of levels $[0, i]$
+and in none of levels $[i + 1, \infty]$. Upon insertion, the new node's level is assigned at random
+according to a geometric distribution; deletion simply removes the node.
+
 \section{Fine-grained Locking Heaps}

### 2014-03-04 paper/paper.tex
Additional notes

@@
 
+% TODO: Mention what we don't cover: non-general PQs, unavailable primitives, ...
 % TODO: In section x, we ... Again, this partially duplicates the last paragraph.
@@
 operations.
-\emph{Quiescently consistent} \cite{shavit1996diffracting}
+\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
 guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
@@
 
-Priority queues are an abstract data structure allowing insertion an items with a given priority
+Priority queues are an abstract data structure allowing insertion of items with a given priority
 and removal of the highest-priority item in logarithmic time. Search trees and Heaps (which are 
-flattened representations complete trees such that each node's value is at least as large as
+flattened representations of complete trees such that each node's value is at least as large as
 those of both children) are concrete data structures which are usually used to implement sequential
@@
 they are both simple and exhibit excellent disjoint-access parallelism. In contrast to
-both search trees and Heaps, reorganization is not necessary since SkipLists rely on
+search trees and Heaps, reorganization is not necessary, since SkipLists rely on
 randomization for an expected $O(\log n)$ complexity of \lstinline|Insert|, \lstinline|Search|
@@
 \section{Fine-grained Locking Heaps}
-\section{SkipList-based Priority Queues}
+
+We chose the priority queue by \citeauthor{hunt1996efficient} \cite{hunt1996efficient}
+as a representative of early concurrent priority queues since it has been proven to
+perform well \cite{shavit2000skiplist} in comparison to other efforts of the time such as \cite{nageshwara1988concurrent,ayani1990lr,yan1998lock}. % Many others \cite{shavit2000skiplist}.
+It is based on a Heap
+structure and attempts to minimize lock contention between threads by a) adding per-node
+locks, b) spreading subsequent insertions through a bit-reversal technique, % Elaborate on these.
+and c) letting insertions traverse bottom-up in order to minimize conflicts with
+top-down deletions.
+
+However, significant limitations to scalability remain. A global lock is required
+to protect accesses to a variable storing the Heap's size which all operations
+must obtain for a short time. Disjoint-access through bit-reversal breaks down
+once a certain amount of traffic is reached, since only subsequent insertions
+are guaranteed to take disjoint paths towards the root node. Note also that
+the root node is a severe serial bottleneck, since it is potentially part of
+every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
+Finally, in contrast to later SkipList-based designs, the capacity of Hunt Heap
+is fixed upon creation.
+
+Benchmarking results in the literature have been mixed; a sequential priority
+queue protected by a single global lock outperforms the \citeauthor{hunt1996efficient}
+Heap in most cases \cite{hunt1996efficient,sundell2003fast}. Speed-up only occurs once
+the size of the Heap reaches a certain threshold such that concurrency
+can be properly exploited (instead of being dominated by global locking overhead).
+
+\section{Lock-free Priority Queues}
+
+Traditional data structures such as the Heap have fallen out of favor;
+instead, SkipLists \cite{pugh1990skip,pugh1998concurrent} have become the focus
+of modern concurrent priority queue research
+\cite{shavit2000skiplist,sundell2003fast,herlihy2012art,linden2013skiplist,alistarhspraylist}.
+SkipLists are both conceptual simple as well as simple to implement; they also exhibit
+excellent disjoint-access properties, and do not require rebalancing due to their
+reliance on randomization for expected $O(\log n)$ time insertion and deletion bounds.
+
+A state of the art lock-free SkipList implementation based on \ac{CAS}
+by \citeauthor{fraser2004practical} \cite{fraser2004practical} is freely available\footnote{
+\url{http://www.cl.cam.ac.uk/research/srg/netos/lock-free/}} under a BSD license.
+\citeauthor{fraser2004practical} exploits unused pointer bits to mark nodes as logically
+deleted, with physical deletion following as a second step.
+
+SkipLists are dynamic data structures in the sense that they grow and shrink
+at runtime. In consequence, careful handling of memory accesses and (de)allocations
+are required. As an additional requirement, these memory management schemes must
+be both scalable and lock-free to avoid limiting the SkipList itself.
+\citeauthor{fraser2004practical} in particular employs lock-free epoch-based garbage-collection.
+
+\subsection{Shavit \& Lotan}
+
+\citeauthor{shavit2000skiplist} were the first to propose the use of SkipLists
+for priority queues \cite{linden2013skiplist}. Their initial implementation
+\cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
+SkipList idea \cite{pugh1998concurrent} which uses one lock per node per level.
+\citeauthor{herlihy2012art} \cite{herlihy2012art} later succeeded in making this
+priority queue lock-free (but not linearizable) and provide a Java implementation
+which we will examine further in the following.
+
+% TODO
+
+\subsection{Sundell \& Tsigas}
+
+\subsection{Linden \& Jonsson}
+
+\section{Relaxed Priority Queues}
+
+\subsection{Wimmer}
+\subsection{SprayList}
+
+\section{Performance Results}
+
+In this section, we compare the performance of several different concurrent priority
+queue implementations:
+
+\begin{itemize}
+\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
+      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
+      However, as the original code was written for the MIPS architecture, we chose to
+      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
+      instead.
+\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
+      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
+      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
+      According to the authors, the commercial Noble library includes a more efficient
+      version of this data structure that can also handle duplicate priority values.
+\item \textbf{Linden} Original code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
+      is available under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
+      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
+      SkipList implementation.
+\end{itemize}
+
+In each test run, the examined priority queue was initially filled with $2^15$
+elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
+for a total of $10$ seconds, where all \lstinline|Insert|
+operations within this context implicitly choose a key uniformly at random from
+the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
+and we report on the average throughput.
+
+All benchmarks were compiled with \verb|-O3| on GCC 4.8.2. Evaluations took place
+on an 80-core Intel Xeon machine.
+
+% TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
 
@@
 * Maybe benchmarks on mars.
+
+Additional notes:
+* Sketch development: Heap, Fraser Skiplist (Details, mention why linearizable),
+  Sundell/Tsigas, other Skiplist-based PQs
+* Modern developments: SprayList, Wimmer et al. Relation to other relaxed structures such as k-FIFO.
 \end{comment}
@@
 \section{Related Work}
+
+% Bounded priority queues
+% PQ's with non-available primitives
+% Other relaxed data structures -> k-FIFO
+% ... ?
 
@@
 
-% TODO: Remove me.
-\nocite{*}
-
 \printbibliography

### 2014-03-04 src/noble.cpp
Free remaining elements in noble PQ

@@ -8,6 +8,12 @@ Noble::Noble()
 
 Noble::~Noble()
 {
+    bool nonempty;
+    uint32_t v;
+    do {
+        nonempty = delete_min(v);
+    } while (nonempty);
+
     delete m_q;
 }
 

### 2014-03-04 src/noble.cpp
Update noble memory management notes

@@ -27,9 +27,9 @@ Noble::insert(const uint32_t v)
 bool
 Noble::delete_min(uint32_t &v)
 {
-    /* TODO: Noble is based on pointers.
-     * Think of some way to avoid losing performance opposed
-     * to other implementations. */
+    /* We tried allocating the ints in advance, but this actually
+     * resulted in lower throughput. Instead, simply alloc/dealloc
+     * in each operation. */
 
     uint32_t *u;
     u = m_q->DeleteMin(nullptr);

### 2014-03-05 paper/paper.tex
Hunt heaps, concurrent SkipLists

@@
 operations.
-\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
+\emph{Quiescently consistent} \cite{shavit1996diffracting}
 guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
@@
 Traditional data structures such as the Heap have fallen out of favor;
-instead, SkipLists \cite{pugh1990skip,pugh1998concurrent} have become the focus
-of modern concurrent priority queue research
+instead, SkipLists \cite{pugh1990skip} have become the focus of modern concurrent
+priority queue research
 \cite{shavit2000skiplist,sundell2003fast,herlihy2012art,linden2013skiplist,alistarhspraylist}.
@@
 
-\subsection{Shavit \& Lotan}
-
-\citeauthor{shavit2000skiplist} were the first to propose the use of SkipLists
-for priority queues \cite{linden2013skiplist}. Their initial implementation
-\cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
-SkipList idea \cite{pugh1998concurrent} which uses one lock per node per level.
-\citeauthor{herlihy2012art} \cite{herlihy2012art} later succeeded in making this
-priority queue lock-free (but not linearizable) and provide a Java implementation
-which we will examine further in the following.
-
-% TODO
+\subsection{Lotan \& Shavit}
 
 \subsection{Sundell \& Tsigas}
-
 \subsection{Linden \& Jonsson}
@@
 \subsection{SprayList}
-
-\section{Performance Results}
-
-In this section, we compare the performance of several different concurrent priority
-queue implementations:
-
-\begin{itemize}
-\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
-      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
-      However, as the original code was written for the MIPS architecture, we chose to
-      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
-      instead.
-\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
-      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
-      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
-      According to the authors, the commercial Noble library includes a more efficient
-      version of this data structure that can also handle duplicate priority values.
-\item \textbf{Linden} Original code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
-      is available under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
-      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
-      SkipList implementation.
-\end{itemize}
-
-In each test run, the examined priority queue was initially filled with $2^15$
-elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
-for a total of $10$ seconds, where all \lstinline|Insert|
-operations within this context implicitly choose a key uniformly at random from
-the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
-and we report on the average throughput.
-
-All benchmarks were compiled with \verb|-O3| on GCC 4.8.2. Evaluations took place
-on an 80-core Intel Xeon machine.
-
-% TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
 
@@
 * Maybe benchmarks on mars.
-
-Additional notes:
-* Sketch development: Heap, Fraser Skiplist (Details, mention why linearizable),
-  Sundell/Tsigas, other Skiplist-based PQs
-* Modern developments: SprayList, Wimmer et al. Relation to other relaxed structures such as k-FIFO.
 \end{comment}
@@
 \section{Related Work}
-
-% Bounded priority queues
-% PQ's with non-available primitives
-% Other relaxed data structures -> k-FIFO
-% ... ?
 
@@
 
+% TODO: Remove me.
+\nocite{*}
+
 \printbibliography

### 2014-03-05 src/pqbench.cpp
Use hwloc to pin to appropriate cores

@@ -1,6 +1,7 @@
 #include <atomic>
 #include <cstdio>
 #include <limits>
+#include <hwloc.h>
 #include <random>
 
 #include "heap.h"
@@ -27,6 +28,8 @@ typedef bool (*fn_delete_min)(uint32_t &);
 
 static fn_insert ins;
 static fn_delete_min del;
+
+static hwloc_topology_t topology;
 
 template <typename T>
 static void
@@ -67,6 +70,24 @@ usage(FILE *out,
         DEFAULT_SIZE);
 }
 
+static void
+pin_to_core(const int id)
+{
+    const int depth = hwloc_get_type_or_below_depth(topology, HWLOC_OBJ_CORE);
+    const int ncores = hwloc_get_nbobjs_by_depth(topology, depth);
+
+    const hwloc_obj_t obj = hwloc_get_obj_by_depth(topology, depth, id % ncores);
+
+    hwloc_cpuset_t cpuset = hwloc_bitmap_dup(obj->cpuset);
+    hwloc_bitmap_singlify(cpuset);
+
+    if (hwloc_set_cpubind(topology, cpuset, HWLOC_CPUBIND_THREAD) != 0) {
+        fprintf(stderr, "Could not bind to core: %s\n", strerror(errno));
+    }
+
+    hwloc_bitmap_free(cpuset);
+}
+
 static void *
 run(void *args)
 {
@@ -77,8 +98,7 @@ run(void *args)
     std::uniform_int_distribution<> rand_int;
     std::uniform_int_distribution<> rand_bool(0, 1);
 
-    /* TODO: Adjust to our machines. */
-    pin(gettid(), as->id);
+    pin_to_core(as->id);
 
     // call in to main thread
     std::atomic_fetch_add(&wait_barrier, 1);
@@ -153,6 +173,9 @@ main(int argc __attribute__ ((unused)),
         exit(EXIT_FAILURE);
     }
 
+    hwloc_topology_init(&topology);
+    hwloc_topology_load(topology);
+
     thread_args_t *ts = new thread_args_t[nthreads];
     memset(ts, 0, nthreads * sizeof(thread_args_t));
 
@@ -195,6 +218,7 @@ main(int argc __attribute__ ((unused)),
     printf("Min ops/t:\t%d\n", min);
     printf("Max ops/t:\t%d\n", max);
 
+    hwloc_topology_destroy(topology);
     delete[] ts;
 
     return 0;

### 2014-03-07 bench.py
Initial benching script

@@ -1,0 +1,62 @@
+#!/usr/bin/env python2
+
+import re
+import subprocess
+
+from optparse import OptionParser
+
+ALGORITHMS = [ 'heap'
+             , 'noble'
+             , 'linden'
+             ]
+
+NCPUS = [  1,  2,  3
+        ,  4,  6,  8, 10
+        , 11, 15, 20
+        , 21, 25, 30
+        , 31, 35, 40
+        , 41, 45, 50
+        , 51, 55, 60
+        , 61, 65, 70
+        , 71, 75, 80
+        ]
+
+BIN = 'build/src/pqbench'
+
+def bench(algorithm, ncpus):
+    output = subprocess.check_output([ BIN
+                                     , '-q', algorithm
+                                     , '-n', str(ncpus)
+                                     ]) # TODO: Size, offset
+
+    print output
+
+if __name__ == '__main__':
+    parser = OptionParser()
+    parser.add_option("-a", "--algorithms", dest = "algorithms",
+            help = "Comma-separated list of ['heap', 'noble', 'linden']")
+    parser.add_option("-n", "--ncpus", dest = "ncpus",
+            help = "Comma-separated list of cpu counts")
+    (options, args) = parser.parse_args()
+
+    if options.algorithms is None:
+        algorithms = ALGORITHMS
+    else:
+        algorithms = options.algorithms.split(',')
+        for a in algorithms:
+            if a not in ALGORITHMS:
+                parser.error('Invalid algorithm')
+
+    if options.ncpus is None:
+        ncpus = NCPUS
+    else:
+        ncpus = list()
+        for n in options.ncpus.split(','):
+            try:
+                ncpus.append(int(n))
+            except:
+                parser.error('Invalid cpu count')
+
+    for a in algorithms:
+        for n in ncpus:
+            bench(a, n)

### 2014-03-07 src/pqbench.cpp
Add verbosity option to benching framework

@@ -15,6 +15,7 @@
 #define DEFAULT_NTHREADS (1)
 #define DEFAULT_OFFSET   (32)
 #define DEFAULT_SIZE     (1 << 15)
+#define DEFAULT_VERBOSE  (false)
 
 static std::atomic<bool> loop;
 static std::atomic<int> wait_barrier;
@@ -68,6 +69,8 @@ usage(FILE *out,
     fprintf(out, "\t-s SIZE\t\tInitialize queue with SIZE elements. "
         "Default: %i\n",
         DEFAULT_SIZE);
+    fprintf(out, "\t-v\tEnable verbose output. Default: %i\n",
+        DEFAULT_VERBOSE);
 }
 
 static void
@@ -134,6 +137,7 @@ main(int argc __attribute__ ((unused)),
     int offset    = DEFAULT_OFFSET;
     int secs      = DEFAULT_SECS;
     int init_size = DEFAULT_SIZE;
+    bool verbose  = DEFAULT_VERBOSE;
 
     const char *type_str = nullptr;
 
@@ -141,7 +145,7 @@ main(int argc __attribute__ ((unused)),
     pq_linden.insert(42);
 
     int opt;
-    while ((opt = getopt(argc, argv, "hn:o:q:s:t:")) >= 0) {
+    while ((opt = getopt(argc, argv, "hn:o:q:s:t:v")) >= 0) {
         switch (opt) {
         case 'h': usage(stdout, argv[0]); exit(EXIT_SUCCESS); break;
         case 'n': nthreads  = atoi(optarg); break;
@@ -149,6 +153,7 @@ main(int argc __attribute__ ((unused)),
         case 'q': type_str  = optarg; break;
         case 's': init_size = atoi(optarg); break;
         case 't': secs      = atoi(optarg); break;
+        case 'v': verbose   = true; break;
         default: assert(0);
         }
     }
@@ -212,11 +217,15 @@ main(int argc __attribute__ ((unused)),
     struct timespec elapsed = timediff(start, end);
     double dt = elapsed.tv_sec + (double)elapsed.tv_nsec / 1000000000.0;
 
-    printf("Total time:\t%1.8f s\n", dt);
-    printf("Ops:\t\t%d\n", sum);
-    printf("Ops/s:\t\t%.0f\n", (double) sum / dt);
-    printf("Min ops/t:\t%d\n", min);
-    printf("Max ops/t:\t%d\n", max);
+    if (verbose) {
[…]

### 2014-03-07 src/pqbench.cpp
Default to an offset of 128 and initial size 2^20

@@ -13,8 +13,8 @@
 
 #define DEFAULT_SECS     (10)
 #define DEFAULT_NTHREADS (1)
-#define DEFAULT_OFFSET   (32)
-#define DEFAULT_SIZE     (1 << 15)
+#define DEFAULT_OFFSET   (128)
+#define DEFAULT_SIZE     (1 << 20)
 #define DEFAULT_VERBOSE  (false)
 
 static std::atomic<bool> loop;

### 2014-03-07 src/pqbench.cpp
Remove unused offset argument

@@ -59,10 +59,6 @@ usage(FILE *out,
     fprintf(out, "\t-t SECS\t\tRun for SECS seconds. "
         "Default: %i\n",
         DEFAULT_SECS);
-    fprintf(out, "\t-o OFFSET\tUse an offset of OFFSET nodes. Sensible "
-        "\n\t\t\tvalues could be 16 for 8 threads, 128 for 32 threads. "
-        "\n\t\t\tDefault: %i\n",
-        DEFAULT_OFFSET);
     fprintf(out, "\t-n NUM\t\tUse NUM threads. "
         "Default: %i\n",
         DEFAULT_NTHREADS);
@@ -134,7 +130,6 @@ main(int argc __attribute__ ((unused)),
      char **argv __attribute__ ((unused)))
 {
     int nthreads  = DEFAULT_NTHREADS;
-    int offset    = DEFAULT_OFFSET;
     int secs      = DEFAULT_SECS;
     int init_size = DEFAULT_SIZE;
     bool verbose  = DEFAULT_VERBOSE;
@@ -149,7 +144,6 @@ main(int argc __attribute__ ((unused)),
         switch (opt) {
         case 'h': usage(stdout, argv[0]); exit(EXIT_SUCCESS); break;
         case 'n': nthreads  = atoi(optarg); break;
-        case 'o': offset    = atoi(optarg); break;
         case 'q': type_str  = optarg; break;
         case 's': init_size = atoi(optarg); break;
         case 't': secs      = atoi(optarg); break;

### 2014-03-07 bench.py
Optionally write results to file

@@ -23,13 +23,16 @@ NCPUS = [  1,  2,  3
 
 BIN = 'build/src/pqbench'
 
-def bench(algorithm, ncpus):
+def bench(algorithm, ncpus, outfile):
     output = subprocess.check_output([ BIN
                                      , '-q', algorithm
                                      , '-n', str(ncpus)
                                      ]) # TODO: Size, offset
 
-    print output
+    outstr = '%s, %d, %s' % (algorithm, ncpus, output.strip())
+
+    print outstr
+    f.write(outstr + '\n')
 
 if __name__ == '__main__':
     parser = OptionParser()
@@ -37,6 +40,8 @@ if __name__ == '__main__':
             help = "Comma-separated list of ['heap', 'noble', 'linden']")
     parser.add_option("-n", "--ncpus", dest = "ncpus",
             help = "Comma-separated list of cpu counts")
+    parser.add_option("-o", "--outfile", dest = "outfile",
+            help = "Write results to outfile")
     (options, args) = parser.parse_args()
 
     if options.algorithms is None:
@@ -57,6 +62,10 @@ if __name__ == '__main__':
             except:
                 parser.error('Invalid cpu count')
 
-    for a in algorithms:
-        for n in ncpus:
-            bench(a, n)
+    if options.outfile is None:
+        options.outfile = '/dev/null'
+
+    with open(options.outfile, 'a') as f:
+        for a in algorithms:
+            for n in ncpus:
+                bench(a, n, f)

### 2014-03-07 bench.py
Use optparse's default option

@@ -36,34 +36,25 @@ def bench(algorithm, ncpus, outfile):
 
 if __name__ == '__main__':
     parser = OptionParser()
-    parser.add_option("-a", "--algorithms", dest = "algorithms",
+    parser.add_option("-a", "--algorithms", dest = "algorithms", default = ",".join(ALGORITHMS),
             help = "Comma-separated list of ['heap', 'noble', 'linden']")
-    parser.add_option("-n", "--ncpus", dest = "ncpus",
+    parser.add_option("-n", "--ncpus", dest = "ncpus", default = ",".join(map(str, NCPUS)),
             help = "Comma-separated list of cpu counts")
-    parser.add_option("-o", "--outfile", dest = "outfile",
+    parser.add_option("-o", "--outfile", dest = "outfile", default = '/dev/null',
             help = "Write results to outfile")
     (options, args) = parser.parse_args()
 
-    if options.algorithms is None:
-        algorithms = ALGORITHMS
-    else:
-        algorithms = options.algorithms.split(',')
-        for a in algorithms:
-            if a not in ALGORITHMS:
-                parser.error('Invalid algorithm')
+    algorithms = options.algorithms.split(',')
+    for a in algorithms:
+        if a not in ALGORITHMS:
+            parser.error('Invalid algorithm')
 
-    if options.ncpus is None:
-        ncpus = NCPUS
-    else:
-        ncpus = list()
-        for n in options.ncpus.split(','):
-            try:
-                ncpus.append(int(n))
-            except:
-                parser.error('Invalid cpu count')
-
-    if options.outfile is None:
-        options.outfile = '/dev/null'
+    ncpus = list()
+    for n in options.ncpus.split(','):
+        try:
+            ncpus.append(int(n))
+        except:
+            parser.error('Invalid cpu count')
 
     with open(options.outfile, 'a') as f:
         for a in algorithms:

### 2014-03-07 bench.py
Add a repetitions argument

@@ -21,6 +21,8 @@ NCPUS = [  1,  2,  3
         , 71, 75, 80
         ]
 
+REPS = 10
+
 BIN = 'build/src/pqbench'
 
 def bench(algorithm, ncpus, outfile):
@@ -42,6 +44,8 @@ if __name__ == '__main__':
             help = "Comma-separated list of cpu counts")
     parser.add_option("-o", "--outfile", dest = "outfile", default = '/dev/null',
             help = "Write results to outfile")
+    parser.add_option("-r", "--reps", dest = "reps", type = 'int', default = REPS,
+            help = "Repetitions per run")
     (options, args) = parser.parse_args()
 
     algorithms = options.algorithms.split(',')
@@ -59,4 +63,5 @@ if __name__ == '__main__':
     with open(options.outfile, 'a') as f:
         for a in algorithms:
             for n in ncpus:
-                bench(a, n, f)
+                for r in xrange(options.reps):
+                    bench(a, n, f)

### 2014-03-10 paper/paper.tex
\nocite{*} is not needed anymore
Further work on evaluation, L&S sections
Benching note
Hardware description TODO
Minor further work

--- 2014-03-04 paper/paper.tex
@@
 
+% TODO: Maybe integrate into description - otherwise remove. Do we need further detail?
+
+\begin{lstlisting}
+struct slist_t {
+  size_t max_level;
+  node_t head[max_level];
+};
+
+struct node_t {
+  key_t key;
+  value_t value;
+  size_t level;
+  node_t next[level];
+};
+
+/* All operations are expected O(log n) time. */
+void slist_insert(slist_t *l, key_t k, value_t v);
+bool slist_delete(slist_t *l, key_t k, value_t &v);
+\end{lstlisting}
+
 \section{Fine-grained Locking Heaps}
@@
 \cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
-SkipList idea \cite{pugh1998concurrent} which uses one lock per node per level.
-\citeauthor{herlihy2012art} \cite{herlihy2012art} later succeeded in making this
-priority queue lock-free (but not linearizable) and provide a Java implementation
-which we will examine further in the following.
+SkipList \cite{pugh1998concurrent} which uses one lock per node per level.
 
-% TODO
+A crucial observation is that nodes nodes which are only partially connected
+do not affect correctness of the data structure. As soon as the first level (i.e. \lstinline|node.level[0]|)
+has been successfully connected, a node is considered to be in the SkipList.
+Therefore, both insertions and deletions can proceed in steps - insertions proceed bottom-up while
+deletions proceed top-down, holding locks only for the current level.
+
+\citeauthor{herlihy2012art} \cite{herlihy2012art} later described a variant of this idea 
+that is lock-free and quiescently consistent (but not linearizable).
 
@@
 % TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
+% Better hardware description - see SprayList section 4.
+% Note: Strong degradation of noble PQ with growing initial size.
 

### 2014-03-11 paper/paper.tex
Shavit & Lotan & Herlihy

@@
 has been successfully connected, a node is considered to be in the SkipList.
-Therefore, both insertions and deletions can proceed in steps - insertions proceed bottom-up while
-deletions proceed top-down, holding locks only for the current level.
+Therefore, both insertions and deletions can be split into steps --- insertions
+proceed bottom-up while deletions proceed top-down, holding locks only for the current level.
 
-\citeauthor{herlihy2012art} \cite{herlihy2012art} later described a variant of this idea 
-that is lock-free and quiescently consistent (but not linearizable).
+\begin{lstlisting}
+struct node_t {
+  [...] /**< Standard node members as above. */
+  atomic<bool> deleted; /**< Initially false. */
+  time_t timestamp;
+  lock_t locks[level + 1];
+};
+\end{lstlisting}
+
+Likewise, deletions are split into a logical phase, atomically setting the \lstinline|node.deleted|
+flag; and a physical phase performs the actual pointer manipulations and can be seen as a simple
+call to the underlying SkipList's \lstinline|sl_delete| function.
+
+A \lstinline|DeleteMin| call starts the list head, and attempts to atomically set
+the deletion flag using a \lstinline|CAS(node.deleted, false, true)| call (or equivalent constructs).
+If it succeeds, the current node is physically deleted and returned to the caller. Otherwise,
+\lstinline|node.next[0]| is set as the new current node and the procedure is repeated.
+If the end of the list is reached, \lstinline|DeleteMin| returns false to indicate an empty list.
+
+Note that so far, this implementation is not linearizable: consider the case in which a
+slow thread A is in the middle of a \lstinline|DeleteMin| call. Within this context, we refer to 
+the node with key $i$ as ``node $i$'', or simply $i$. Several \lstinline|CAS|
+operations have failed, and A is currently at node $j$.
+A fast thread B then first inserts a node $i$, followed by a node $k$ such that
+$i < j < k$, i.e. the former and latter nodes are inserted, respectively, before and after
+thread A's current node. Assuming further that all nodes between $j$ and $k$ have already
+been deleted, then thread A will return node $k$. This execution is not linearizable; however,
+it is quiescently consistent since operations can be reordered to correspond to some sequential
+execution at quiescent periods.
+
+\citeauthor{shavit2000skiplist} counteract this by introducing a \lstinline|timestamp| for each
+node which is set upon successful insertion. Each \lstinline|DeleteMin| operation 
+simply ignores all nodes it sees that have not been fully inserted at the time it was called.
+
+Explicit memory management is required to avoid dereferencing pointers to freed memory areas
+by other threads after physical deletion. This implementation uses a dedicated garbage collector
+thread in combination with a timestamping mechanism which frees \lstinline|node|'s memory only
+when all threads that might have seen a pointer to \lstinline|node| have exited the data structure.
+
+\citeauthor{herlihy2012art} \cite{herlihy2012art} later described and implemented a variant of this idea 
+in Java that is lock-free and quiescently consistent. While mostly identical, notable differences are
+that a) the new variant is based on a lock-free skiplist, b) the timestamping mechanism was not
+employed and thus linearizability was lost, and c) explicit memory management is not required
+because the Java virtual machine provides a garbage collector.
 

### 2014-03-11 paper/paper.tex
Sundell & Tsigas section

@@
 
+\begin{figure}[h]
 \begin{lstlisting}
@@
   size_t level;
-  node_t next[level];
+  node_t *next[level];
 };
@@
 \end{lstlisting}
+\caption{Basic SkipList structure and operations.}
+\label{fig:basicsl}
+\end{figure}
 
-\section{Fine-grained Locking Heaps}
+\section{Fine-grained Locking Heaps} \label{sec:hunt}
 
@@
 
-\subsection{Shavit \& Lotan}
+\subsection{Shavit \& Lotan} \label{sec:shavit}
 
@@
 SkipList \cite{pugh1998concurrent} which uses one lock per node per level.
+
+% TODO: This counts for all skiplist impls.
 
@@
 \end{lstlisting}
+\caption{\citeauthor{shavit2000skiplist} structure.}
+\label{fig:slsl}
+\end{figure}
 
@@
 
+% TODO: Linearizability points.
+
 Explicit memory management is required to avoid dereferencing pointers to freed memory areas
@@
 
-\subsection{Sundell \& Tsigas}
+\subsection{Sundell \& Tsigas} \label{sec:sundell}
+
+\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
+\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
+and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
+In contrast to other structures covered in this paper, this priority queue is restricted to
+contain items with pairwise distinct priorities.
+A real-time version is also provided which will we will not discuss further (interested readers are
+referred to \cite{sundell2003fast}).
+
+\begin{figure}[h]
+\begin{lstlisting}
+struct node_t {
+  [...] /**< Standard node members as above. */
+  size_t valid_level;
+  node_t *prev;
+};
+\end{lstlisting}
+\caption{\citeauthor{sundell2003fast} structure.}
+\label{fig:slsl}
+\end{figure}
+
+The structure of each node is basically identical to Figure \ref{fig:basicsl}. However, \citeauthor{sundell2003fast}
+exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
+are unused and reuse pointers as deletion marks. A set least significant bit signifies on a pointer signifies
+that the current node is about to be deleted. We will use the notation 
+\lstinline|CAS(node.level[i], (ptr, false), (ptr, true))| to signify \ac{CAS} operations on the
+combined pointer \& flag fields.
+Reuse of \lstinline|node.level[i]| pointers
+prevents situations in which a new node is inserted while its predecessor is being removed,
+effectively deleting both nodes from the list. Likewise, the reuse of the \lstinline|node.value|
+pointer ensures that updates of pointer values (which occur when a node with the inserted priority already exists)
+handle concurrent node removals correctly.
+
+As in the \citeauthor{shavit2000skiplist} priority queue, insertions proceed bottom-up while
+deletions proceed top-down --- on the one hand, the choice of opposite directions reduces collisions
+between concurrent insert and delete operations, while on the other hand removing nodes from top levels first
+allows many other concurrent operations to simply skip these nodes, further improving performance.
+\lstinline|node.valid_level| is updated during inserts to always equal the highest level of the SkipList
+at which pointers in this node have already been set (as opposed to \lstinline|node.level|, which equals
+the final level of the node).
+
+A helping mechanism is employed whenever a node is encountered that has its deletion bit set, which attempts
+to set the deletion bits on all next pointers and then removes the node from the current level. The
+\lstinline|node.prev| pointer is used as a shortcut to the previous node, avoiding a complete retraversal
+of the list.
+
+This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
+\cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
+\cite{michael1995correction}. It was chosen in particular because this scheme can guarantee validity
+of \lstinline|prev| as well as all \lstinline|next| pointers. Additionally, it does not require a separate
+garbage collector thread.
+
+A rigorous linearizability proof is provided in the original paper \cite{sundell2003fast} which shows
+linearizability points for all possible outcomes of both \lstinline|Insert| and \lstinline|DeleteMin|
+operations.
+
+% TODO: Further details?
+
+Benchmarks performed by \citeauthor{sundell2003fast} show their queue performing noticeably better than both locking 
+queues from Sections \ref{sec:shavit} and \ref{sec:hunt}, and slightly better than a priority queue
+consisting of a SkipList protected by a single global lock.
 
@@
 
-In each test run, the examined priority queue was initially filled with $2^15$
+In each test run, the examined priority queue was initially filled with $2^{15}$
 elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions

### 2014-03-13 paper/paper.tex
Minor corrections & additions
TODO note
Notes and relaxed introduction

@@
 \caption{\citeauthor{shavit2000skiplist} structure.}
-\label{fig:slsl}
+\label{fig:shavitsl}
 \end{figure}
@@
 \caption{\citeauthor{sundell2003fast} structure.}
-\label{fig:slsl}
+\label{fig:sundellsl}
 \end{figure}
@@
 
-\subsection{Linden \& Jonsson}
+\subsection{Linden \& Jonsson} \label{sec:linden}
+
+\begin{figure}[h]
+\begin{lstlisting}
+struct node_t {
+  [...] /**< Standard node members as above. */
+  atomic<bool> inserting;
+};
+\end{lstlisting}
+\caption{\citeauthor{linden2013skiplist} structure.}
+\label{fig:lindensl}
+\end{figure}
+
+ % Mention the BoundOffset parameter (p17), and benching using Fraser-based impls.
 
@@
 
-\subsection{Wimmer}
-\subsection{SprayList}
+The body of work discussed in previous sections
+creates the impression that limits of strict, deterministic priority queues have been reached.
+In particular, \citeauthor{} conclude that scalability is solely limited by \lstinline|DeleteMin|,
+and that less than one modified memory location per thread and operation would have to be read
+in order to achieve improved performance \cite{}. % TODO: Cites for this entire section.
+
+Recently, relaxation of provided guarantees have been investigated as another method of reducing
+contention and improving disjoint-access parallelism.
+For instance, k-FIFO queues \cite{} have achieved considerable
+speed-ups compared to strict FIFO queues by allowing \lstinline|Dequeue| to return elements
+up to k positions out of order.
+
+Relaxation has also been applied to concurrent priority queues with some success, and in the following
+sections we discuss two different approaches.
+
+\subsection{Wimmer} \label{sec:wimmer}
+\subsection{SprayList} \label{sec:spraylist}
 
@@
 % TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
-% Better hardware description - see SprayList section 4.
+% Better hardware description - see SprayList section 4, Linden p16.
 % Note: Strong degradation of noble PQ with growing initial size.
+% Standard PQ benchmarking methodology, cite linden + ??.
+% Drop-off when using more than 1 socket, Linden p17.
 

### 2014-03-13 paper/paper.tex
Initial Wimmer paragraphs

@@
 creates the impression that limits of strict, deterministic priority queues have been reached.
-In particular, \citeauthor{} conclude that scalability is solely limited by \lstinline|DeleteMin|,
+In particular, \citeauthor{linden2013skiplist} conclude that scalability is solely limited by \lstinline|DeleteMin|,
 and that less than one modified memory location per thread and operation would have to be read
-in order to achieve improved performance \cite{}. % TODO: Cites for this entire section.
+in order to achieve improved performance \cite{linden2013skiplist}. % TODO: Cites for this entire section.
 
@@
 Relaxation has also been applied to concurrent priority queues with some success, and in the following
-sections we discuss two different approaches.
+sections we discuss two such approaches.
 
 \subsection{Wimmer} \label{sec:wimmer}
+
+\citeauthor{wimmer2013data} presented the first relaxed, linearizable, and lock-free priority queue
+in \cite{wimmer2013data}. An open-source implementation is available in the task-scheduling system
+\emph{Pheet}\footnote{\url{http://pheet.org}}.
+
+The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
+of Dijkstra's shortest path algorithm \cite{}. Their model creates a task for each node
+expansion, and therefore comes with a considerable task scheduling overhead. Nonetheless,
+the parallel implementation scales well up to 10 threads, and further limited performance
+gains are made until 40 threads. To date, no direct comparisons to other concurrent
+priority queues have been possible since the data structure is strongly tied to \emph{Pheet},
+but a separate implementation of k-priority queues is planned.
+
 \subsection{SprayList} \label{sec:spraylist}

### 2014-03-13 paper/paper.tex
Initial Linden paragraphs

@@
 
-\title{Lock-free Skiplist-based Priority Queues \\
+\title{State of the Art in Concurrent Priority Queues \\
        Student Paper for Seminar in Algorithms 2013W \\
@@
 
+% Logical/physical deletion separation allows concurrent physical deletions.
+
 Note that so far, this implementation is not linearizable: consider the case in which a
@@
 
+% TODO: Mention backoff.
+
 This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
@@
 
+One of the most recent priority queue implementations was published by \citeauthor{linden2013skiplist}
+in \citeyear{linden2013skiplist} \cite{linden2013skiplist}. They present a linearizable, lock-free concurrent priority
+queue which achieves a speed-up of $30-80\%$ compared to other SkipList-based concurrent priority queues by
+minimizing the number of \ac{CAS} operations within most \lstinline|DeleteMin| operations.
+
+A priority queue implementation is called deterministic when the algorithm does not contain randomized elements.
+It is called strict when \lstinline|DeleteMin| is guaranteed to return the minimal element currently within the queue
+(as opposed to relaxed data structures which are discussed further in the next section).
+All such priority queues share an inherent bottleneck, since all threads calling \lstinline|DeleteMin| compete
+for the minimal element, causing both contention through concurrent \ac{CAS} operations on the same variable
+as well as serialization effort by the cache coherence protocol for all other processor accessing the same cache
+line.
+
 \begin{figure}[h]
@@
 \end{figure}
+
+As in Section \ref{sec:sundell}, deletion flags are packed into each node's pointers. However,
+% Deletion flag only on next[0]!
+
+%  ``Entirely limited by logical deletion in DeleteMin''
 

### 2014-03-13 paper/paper.tex
Minor edits in evaluation

@@
       version of this data structure that can also handle duplicate priority values.
-\item \textbf{Linden} Original code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
-      is available under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
+\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
+      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
       It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
-      SkipList implementation.
+      SkipList implementation. The aim of this implementation is to minimize contention in
+      calls to \lstinline|DeleteMin|.
 \end{itemize}

### 2014-03-17 paper/paper.tex
Use \citeauthor in section titles
Finish up the Linden section

@@
 
-\subsection{Shavit \& Lotan} \label{sec:shavit}
+\subsection{\citeauthor{shavit2000skiplist}} \label{sec:shavit}
 
@@
 
-\subsection{Sundell \& Tsigas} \label{sec:sundell}
+\subsection{\citeauthor{sundell2003fast}} \label{sec:sundell}
 
@@
 
-\subsection{Linden \& Jonsson} \label{sec:linden}
+\subsection{\citeauthor{linden2013skiplist}} \label{sec:linden}
 
@@
 
-As in Section \ref{sec:sundell}, deletion flags are packed into each node's pointers. However,
-% Deletion flag only on next[0]!
+In this implementation, \lstinline|DeleteMin| operations only perform logical deletion by setting
+the deletion flag with a single \ac{FAO} call; nodes are only deleted physically once a certain
+threshold of logically deleted nodes is reached.
 
-%  ``Entirely limited by logical deletion in DeleteMin''
+This mechanism requires a new invariant, in that the set of all logically deleted nodes must always
+form a prefix of the SkipList. Recall that in the \citeauthor{sundell2003fast} queue, deletion flags
+for node \lstinline|n| were packed into \lstinline|n.next| pointers, preventing insertion of new
+nodes \emph{after} deleted nodes. This implementation instead places the deletion flag into the
+lowest level \lstinline|next| pointer of the previous node, preventing inserations \emph{before}
+logically deleted nodes.
 
- % Mention the BoundOffset parameter (p17), and benching using Fraser-based impls.
+Once the prefix of logically deleted nodes reaches a specified length \lstinline|BoundOffset|,
+the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
+deletions by updating \lstinline|slist.head[0]| to point at the last logically deleted node with a
+single \ac{FAO} operation. The remaining \lstinline|slist.head| pointers are then updated, and
+all physically deleted nodes are marked for recycling.
+
+Since at any time, the data structure contains a prefix of logically deleted nodes, all \lstinline|DeleteMin|
+operations must traverse this sequence before reaching a non-deleted node. In general, reads of nonmodified
+memory locations are very cheap; however, benchmarks in \cite{linden2013skiplist} have shown that
+after a certain point, the effort spent in long read sequences significantly outweighs the reduced
+number of \ac{CAS} calls. It is therefore crucial to choose \lstinline|BoundOffset| carefully, with the
+authors recommending a prefix length bound of 128 for 32 threads.
+
+The actual \lstinline|DeleteMin| and \lstinline|Insert| implementations are surprisingly simple.
+Deletions simply traverse the list until the first node for which \lstinline|(ptr, d) = FAO((node.next[0], d), 1)|
+returns a previously unset deletion flag (\lstinline|d = 0|) and then return \lstinline|ptr|.
+Insertions occur bottom-up and follow the basic \citeauthor{fraser2004practical} algorithm \cite{fraser2004practical},
+taking the separation of deletion flags and nodes into account.
+
+The authors provide high level correctness and linearizability proofs as well as a model for the
+SPIN model checker. Performance has been shown to compare favorably to both
+\citeauthor{sundell2003fast} and \citeauthor{shavit2000skiplist} queues, with throughput improved by
+up to $80\%$.
 
@@
 
-\subsection{Wimmer} \label{sec:wimmer}
+\subsection{\citeauthor{wimmer2013data}} \label{sec:wimmer}
 

### 2014-03-17 paper/paper.tex
Extend Wimmer section

@@
 
+% TODO: Mention inserting flag and memory management!
+
 In this implementation, \lstinline|DeleteMin| operations only perform logical deletion by setting
@@
 
-Once the prefix of logically deleted nodes reaches a specified length \lstinline|BoundOffset|,
+Once the prefix of logically deleted nodes reaches a specified length {\lstset{breaklines,breakatwhitespace} \lstinline|BoundOffset|},
 the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
@@
 The body of work discussed in previous sections
-creates the impression that limits of strict, deterministic priority queues have been reached.
+creates the impression that the outer limits of strict, deterministic priority queues have been reached.
 In particular, \citeauthor{linden2013skiplist} conclude that scalability is solely limited by \lstinline|DeleteMin|,
 and that less than one modified memory location per thread and operation would have to be read
-in order to achieve improved performance \cite{linden2013skiplist}. % TODO: Cites for this entire section.
+in order to achieve improved performance \cite{linden2013skiplist}.
 
@@
 contention and improving disjoint-access parallelism.
-For instance, k-FIFO queues \cite{} have achieved considerable
-speed-ups compared to strict FIFO queues by allowing \lstinline|Dequeue| to return elements
-up to k positions out of order.
+For instance, k-FIFO queues \cite{kirsch2012fast} have achieved considerable
+speed-ups compared to strict FIFO queues by allowing {\lstset{breaklines,breakatwhitespace} \lstinline|Dequeue|} to return elements
+up to $k$ positions out of order.
 
@@
 \citeauthor{wimmer2013data} presented the first relaxed, linearizable, and lock-free priority queue
-in \cite{wimmer2013data}. An open-source implementation is available in the task-scheduling system
-\emph{Pheet}\footnote{\url{http://pheet.org}}.
+in \cite{wimmer2013data}. It is integrated as a priority scheduler into their \emph{Pheet} task-scheduling
+system, and an open-source implementation is available\footnote{\url{http://pheet.org}}.
+
+Their paper presents several variations on the common theme of priority queues: a distributed work-stealing
+queue which can give no guarantees as to global ordering since it consists of separate priority queues
+at each thread; a relaxed centralized priority queue in which no more than $k$ items are missed
+by any processor; and a relaxed hybrid data structure which combines both ideas and provides
+a guarantee that no thread misses more than $kP$ items where $P$ is the number of participating threads.
+In this section, we examine only the hybrid variant since it provides both the scalability of work-stealing
+queues and the ordering guarantees of the centralized priority queue.
+
+\begin{figure}[h]
+\begin{lstlisting}
+struct globals_t {
+  list_of_item_t global_list;
+};
+
+/* Thread-local items. */
+struct locals_t {
+  list_of_item_t local_list;
+  pq_t prio_queue;
+  size_t remaining_k;
+};
+\end{lstlisting}
+\caption{\citeauthor{wimmer2013data} structure.}
+\label{fig:wimmerq}
+\end{figure}
+
+The hybrid queue consists of a list of globally visible items and one local item list as well as a local
+sequential priority queue per thread. The thread-local counter \lstinline|remaining_k| tracks how many more
+items may be added to the local queue until all local items must be made globally visible to avoid
+breaking guarantees.
+
+Whenever an item is added, it is first added to both \lstinline|local_list| and \lstinline|prio_queue|
+and \lstinline|remaining_k| is decremented. If \lstinline|remaining_k| reaches zero, then the local
+item list is appended to the global list, and all not yet seen (by this thread) items of the global
+queue are added to the local priority queue.
+
+\lstinline|DeleteMin| simply pops the local priority queue as long is it is non-empty and the popped
+item has already been deleted. When a non-deleted item is popped, it is atomically marked as deleted
+and returned to the caller. If instead we are faced with an empty local queue, we attempt to spy,
+i.e. copy items from another thread's local list.
 
 The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
-of Dijkstra's shortest path algorithm \cite{}. Their model creates a task for each node
+of Dijkstra's shortest path algorithms. Their model creates a task for each node
 expansion, and therefore comes with a considerable task scheduling overhead. Nonetheless,

### 2014-03-17 paper/paper.tex
SprayList section, some benchmarking

@@
 
-\section{Lock-free Priority Queues}
+\section{Lock-free Priority Queues} \label{sec:lockfree}
 
@@
 
+The SprayList is another recent approach towards a relaxed priority queue by \citeauthor{alistarhspraylist}
+\cite{alistarhspraylist}. Instead of the distributed approach described in the previous section,
+the SprayList is based on a central data structure, and uses a random walk in \lstinline|DeleteMin|
+in order to spread accesses over the $O(P \log^3 P)$ first elements with high probability, where $P$
+is again the number of participating threads.
+
+\citeauthor{fraser2004practical}'s lock-free SkipList \cite{fraser2004practical} again serves as the
+basis for the priority queue implementation. The full source code is available online\footnote{
+\url{https://github.com/jkopinsky/SprayList}}. In the SprayList, \lstinline|Insert| calls are simply
+forwarded to the underlying SkipList.
+
+The \lstinline|DeleteMin| operation however executes a random walk, also called a \emph{spray}, the
+purpose of which is to spread accesses over a certain section of the SkipList uniformly such that
+collisions between multiple concurrent \lstinline|DeleteMin| calls are unlikely. This is achieved by
+starting at some initial height, walking a randomized number of steps, descending a randomized number of levels,
+and repeating this procedure until a node $n$ is reached on the lowest level. If $n$ is not deleted,
+it is logically deleted and returned. Otherwise, a \emph{spray} is either reattempted, or the thread
+becomes a cleaner, traversing the lowest level of the SkipList and physically removing logically deleted
+nodes it comes across. A number of dummy nodes are added to the beginning of the list in order to counteract
+the algorithm's bias against initial items.
+
+The \emph{spray} parameters are chosen such that with high probability, one of the $O(P \log^3 P)$
+first elements is returned, and that each of these elements is chosen roughly uniformly at random.
+The final effect is that accesses to the data structure are spread out, reducing contention and resulting
+in a noticeably lower number of \ac{CAS} failures in comparison to strict priority queues described
+in Section \ref{sec:lockfree}.
+
+The authors do not provide any statement as to the linearizability (or other concurrent correctness
+criteria) of the SprayList, and it is not completely clear how to define it since no
+sequential semantics are given.
+
+Benchmarks show promising results: the SprayList scales well at least up to 80 threads,
+and performs close (within a constant factor) to an implementation using a random remove instead
+of \lstinline|DeleteMin|, which the authors consider as the performance ideal.
+
+% Maybe mention that P log^3 P is a very loose bound.
+
 \section{Performance Results}
@@
 \begin{itemize}
+\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
+      by the C++ standard library, protected by a single global lock.
 \item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
@@
       calls to \lstinline|DeleteMin|.
+\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
+      SkipList using random walks to spread data accesses
+      incurred by \lstinline|DeleteMin| calls. Code provided by \citeauthor{alistarhspraylist} is
+      available on Github\footnote{\url{https://github.com/jkopinsky/SprayList}}.
 \end{itemize}
@@
 operations within this context implicitly choose a key uniformly at random from
-the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
+the range of all 32-bit integers. This methodology seems to be the de facto standard for concurrent % TODO: Similar phrasing as linden :(
+priority queue benchmarks
+\cite{alistarhspraylist,linden2013skiplist,shavit2000skiplist,sundell2003fast}.
+Each run was repeated for a total of $10$ times
 and we report on the average throughput.
 
-All benchmarks were compiled with \verb|-O3| on GCC 4.8.2. Evaluations took place
-on an 80-core Intel Xeon machine.
+All benchmarks were compiled with \verb|-O3| using GCC 4.8.2. Evaluations took place
+on an 80-core Intel Xeon machine. The benchmarking code was adapted from \citeauthor{linden2013skiplist}'s
+benchmarking suite and is available at \url{https://github.com/schuay/seminar_in_algorithms}.
 

### 2014-03-17 paper/paper.tex
Related work

@@
 
+% TODO: Memory management.
+
 The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
@@
 
+There have been many publications on more specialized variants of concurrent priority queues
+which we did not cover in this paper.
+
+% PQ's with non-available primitives
+
+\citeauthor{liu2012lock} present a lock-free, array-based priority queue in \cite{liu2012lock}
+using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the standard
+\lstinline|Insert| and \lstinline|DeleteMin| operations, it also provides an \lstinline|ExtractMany|
+function which returns the $n$ highest priority items in the queue, and a function that returns
+items which ``probably'' have high priority. Other implementations based on non-standard instructions
+can be found in \cite{israeli1993efficient,greenwald1999non}.
+
 % Bounded priority queues
-% PQ's with non-available primitives
+
+Another possibility is the restriction of priorities to a particular set, i.e. bounded range priorities.
+\citeauthor{shavit1999scalable} present two such data structures in \cite{shavit1999scalable} using
+combining funnels and bins of items.
+
+% PQ's in message passing systems (see SprayList reviews)
+
+Concurrent priority queues have also been studied in other contexts such as distributed memory systems.
+\citeauthor{karp1993randomized} describe a distributed priority queue in which each processor owns
+a local priority queue, inserts are sent to a random priority queue, and \lstinline|DeleteMin| operations
+simply access the local queue \cite{karp1993randomized}.
+\citeauthor{sanders1998randomized} extends this idea to remove the globally best elements instead
+\cite{sanders1998randomized}.
+
 % Other relaxed data structures -> k-FIFO
-% ... ?
+
+Finally, the principle of relaxation has also been applied to other data structures.
+\citeauthor{kirsch2012fast} invented an efficient k-FIFO queue \cite{kirsch2012fast},
+% TODO: More cites.
 

### 2014-03-18 src/globallock.cpp
Add GlobalLock to benchmark

@@ -1,0 +1,25 @@
+#include "globallock.h"
+
+
+bool
+GlobalLock::delete_min(uint32_t &v)
+{
+    std::lock_guard<std::mutex> g(m_mutex);
+
+    if (m_q.empty()) {
+        return false;
+    }
+
+    v = m_q.top();
+    m_q.pop();
+
+    return true;
+}
+
+void
+GlobalLock::insert(const uint32_t v)
+{
+    std::lock_guard<std::mutex> g(m_mutex);
+
+    m_q.push(v);
+}

### 2014-03-18 src/globallock.h

@@ -1,0 +1,22 @@
+#ifndef __GLOBALLOCK_H
+#define __GLOBALLOCK_H
+
+#include <mutex>
+#include <queue>
+
+class GlobalLock
+{
+public:
+    void init_thread(const size_t) { }
+
+    void insert(const uint32_t v);
+    bool delete_min(uint32_t &v);
+
+private:
+    typedef std::priority_queue<uint32_t> pq_t;
+
+    std::mutex m_mutex;
+    pq_t m_q;
+};
+
+#endif /* __GLOBALLOCK_H */

### 2014-03-18 src/pqbench.cpp

@@ -4,6 +4,7 @@
 #include <hwloc.h>
 #include <random>
 
+#include "globallock.h"
 #include "heap.h"
 #include "linden.h"
 #include "noble.h"
@@ -20,6 +21,7 @@
 static std::atomic<bool> loop;
 static std::atomic<int> wait_barrier;
 
+static GlobalLock pq_globallock;
 static Heap pq_heap(DEFAULT_SIZE << 2);
 static Noble pq_noble;
 static Linden pq_linden(DEFAULT_OFFSET);
@@ -55,7 +57,7 @@ usage(FILE *out,
         "Options:\n", argv0);
 
     fprintf(out, "\t-h\t\tDisplay usage.\n");
-    fprintf(out, "\t-q QUEUE\tRun benchmarks on queue of type TYPE (heap|linden|noble).\n");
+    fprintf(out, "\t-q QUEUE\tRun benchmarks on queue of type TYPE (globallock|heap|linden|noble).\n");
     fprintf(out, "\t-t SECS\t\tRun for SECS seconds. "
         "Default: %i\n",
         DEFAULT_SECS);
@@ -155,6 +157,10 @@ main(int argc __attribute__ ((unused)),
     if (type_str == nullptr) {
         usage(stderr, argv[0]);
         exit(EXIT_FAILURE);
+    } else if (strcmp(type_str, "globallock") == 0) {
+        ins = [](const uint32_t v) { pq_globallock.insert(v); };
+        del = [](uint32_t &v) { return pq_globallock.delete_min(v); };
+        pq_init(pq_globallock, init_size);
     } else if (strcmp(type_str, "heap") == 0) {
         ins = [](const uint32_t v) { pq_heap.insert(v); };
         del = [](uint32_t &v) { return pq_heap.delete_min(v); };

### 2014-03-18 lib/spraylist/skiplist.h
Include utils.h in spraylist/skiplist.h
The ALIGNED macro is defined in utils.h.

@@ -37,6 +37,7 @@
 
 #include "tm.h"
 #include "ssalloc.h"
+#include "utils.h"
 
 #define DEFAULT_DURATION                1000
 #define DEFAULT_INITIAL                 1024

### 2014-03-18 lib/spraylist/skiplist.h
Alter SprayList val_t to uint32_t

@@ -60,7 +60,7 @@ extern uint8_t levelmax[64];
 
 #define TRANSACTIONAL                   d->unit_tx
 
-typedef unsigned long val_t;
+typedef uint32_t val_t;
 typedef intptr_t level_t;
 #define VAL_MIN                         INT_MIN
 #define VAL_MAX                         INT_MAX

### 2014-03-18 lib/linden/common.h
Integrate the SprayList

@@ -44,6 +44,7 @@
 typedef struct thread_args_s
 {
     pthread_t thread;
+    size_t nthreads;
     int id;
     gsl_rng *rng;
     int measure;

### 2014-03-18 src/pqbench.cpp

@@ -8,6 +8,7 @@
 #include "heap.h"
 #include "linden.h"
 #include "noble.h"
+#include "spraylist.h"
 
 #undef max /* Clash between macro and limits. */
 #undef min
@@ -25,6 +26,7 @@ static GlobalLock pq_globallock;
 static Heap pq_heap(DEFAULT_SIZE << 2);
 static Noble pq_noble;
 static Linden pq_linden(DEFAULT_OFFSET);
+static SprayList pq_spraylist;
 
 typedef void (*fn_insert)(const uint32_t);
 typedef bool (*fn_delete_min)(uint32_t &);
@@ -57,7 +59,7 @@ usage(FILE *out,
         "Options:\n", argv0);
 
     fprintf(out, "\t-h\t\tDisplay usage.\n");
-    fprintf(out, "\t-q QUEUE\tRun benchmarks on queue of type TYPE (globallock|heap|linden|noble).\n");
+    fprintf(out, "\t-q QUEUE\tRun benchmarks on queue of type TYPE (globallock|heap|linden|noble|spraylist).\n");
     fprintf(out, "\t-t SECS\t\tRun for SECS seconds. "
         "Default: %i\n",
         DEFAULT_SECS);
@@ -98,6 +100,9 @@ run(void *args)
     std::mt19937 gen(rd());
     std::uniform_int_distribution<> rand_int;
     std::uniform_int_distribution<> rand_bool(0, 1);
+
+    /* Special handling for SprayList. */
+    pq_spraylist.init_thread(as->nthreads);
 
     pin_to_core(as->id);
 
@@ -173,6 +178,10 @@ main(int argc __attribute__ ((unused)),
         ins = [](const uint32_t v) { pq_noble.insert(v); };
         del = [](uint32_t &v) { return pq_noble.delete_min(v); };
         pq_init(pq_noble, init_size);
+    } else if (strcmp(type_str, "spraylist") == 0) {
+        ins = [](const uint32_t v) { pq_spraylist.insert(v); };
+        del = [](uint32_t &v) { return pq_spraylist.delete_min(v); };
+        pq_init(pq_spraylist, init_size);
     } else {
         usage(stderr, argv[0]);
         exit(EXIT_FAILURE);
@@ -187,6 +196,7 @@ main(int argc __attribute__ ((unused)),
     thread_args_t *t;
     for (int i = 0; i < nthreads && (t = &ts[i]); i++) {
         t->id = i;
[…]

### 2014-03-18 src/spraylist.cpp

@@ -1,0 +1,62 @@
+#include "spraylist.h"
+
+extern "C" {
+#include "spraylist/include/random.h"
+#include "spraylist/intset.h"
+#include "spraylist/linden.h"
+}
+
+constexpr unsigned int INITIAL_SIZE = 1 << 15;
+
+/** See documentation of --elasticity in spraylist/test.c. */
+#define READ_ADD_REM_ELASTIC_TX (4)
+
+static __thread bool initialized = false;
+static __thread thread_data_t *d;
+
+__thread unsigned long *seeds;
+
+
+SprayList::SprayList()
+{
+    init_thread(1);
+    *levelmax = floor_log_2(INITIAL_SIZE);
+    m_q = sl_set_new();
+}
+
+SprayList::~SprayList()
+{
+    sl_set_delete(m_q);
+    delete d;
+}
+
+void
+SprayList::init_thread(const size_t nthreads)
+{
+    if (!initialized) {
+        seeds = seed_rand();
+
+        ssalloc_init();
+
+        d = new thread_data_t;
+        d->seed = rand();
+        d->seed2 = rand();
+
+        initialized = true;
+    }
+
+    d->nb_threads = nthreads;
+}
+
+void
+SprayList::insert(const uint32_t v)
+{
+    sl_add(m_q, v, READ_ADD_REM_ELASTIC_TX);
+}
+
+bool
+SprayList::delete_min(uint32_t &v)
+{
+    const int ret = spray_delete_min(m_q, &v, d);
+    return (ret == 1);
+}

### 2014-03-18 src/spraylist.h

@@ -1,0 +1,25 @@
+#ifndef __SPRAYLIST_H
+#define __SPRAYLIST_H
+
+extern "C" {
+#include "spraylist/pqueue.h"
+}
+
+class SprayList
+{
+public:
+    SprayList();
+    virtual ~SprayList();
+
+    void init_thread(const size_t nthreads);
+
+    void insert(const uint32_t v);
+    bool delete_min(uint32_t &v);
+
+private:
+    typedef sl_intset_t pq_t;
+
+    pq_t *m_q;
+};
+
+#endif /* __SPRAYLIST_H */

### 2014-03-18 bench.py
Add globallock and spraylist to bench.py

@@ -5,9 +5,11 @@ import subprocess
 
 from optparse import OptionParser
 
-ALGORITHMS = [ 'heap'
+ALGORITHMS = [ 'globallock'
+             , 'heap'
              , 'noble'
              , 'linden'
+             , 'spraylist'
              ]
 
 NCPUS = [  1,  2,  3

### 2014-03-18 src/linden.cpp
Add the Linden queue

@@ -20,13 +20,13 @@ Linden::Linden(const int max_offset)
 Linden::~Linden()
 {
 	pq_destroy(m_q);
+    _destroy_gc_subsystem();
 }
 
 void
 Linden::insert(const uint32_t v)
 {
 	linden_insert(m_q, v);
-	_destroy_gc_subsystem();
 }
 
 bool

### 2014-03-18 lib/libcds/cds/container/mspriority_queue.h
Fix most remaining warnings in external code

@@ -421,7 +421,7 @@ namespace cds { namespace container {
         void clear()
         {
 #       ifdef CDS_CXX11_LAMBDA_SUPPORT
-            clear_with( []( value_type const& src ) {} ) ;
+            clear_with( []( value_type const& ) {} ) ;
 #       else
             clear_with( empty_cleaner() ) ;
 #       endif

### 2014-03-18 lib/linden/common.h

@@ -1,6 +1,6 @@
 #ifndef COMMON_H
 #define COMMON_H
-#define _GNU_SOURCE
+
 #include <inttypes.h>
 #include <unistd.h>
 #include <assert.h>

### 2014-03-18 lib/spraylist/include/random.h

@@ -74,7 +74,7 @@ rand_range(long r)
 
 /* Re-entrant version of rand_range(r) */
 static inline long
-rand_range_re(unsigned int *seed, long r) 
+rand_range_re(unsigned int *seed __attribute__ ((unused)), long r)
 {
   /* PF_START(0); */
 #if defined(LOCAL_RAND)

### 2014-03-18 lib/spraylist/include/utils.h

@@ -222,14 +222,6 @@ extern "C" {
 #  define CORES_PER_SOCKET 8
 #  define CACHE_LINE_SIZE 64
 #  define NOP_DURATION 1
-  static uint8_t  the_cores[] = {
-    0, 1, 2, 3, 4, 5, 6, 7, 
-    8, 9, 10, 11, 12, 13, 14, 15, 
-    16, 17, 18, 19, 20, 21, 22, 23, 
-    24, 25, 26, 27, 28, 29, 30, 31, 
-    32, 33, 34, 35, 36, 37, 38, 39, 
-    40, 41, 42, 43, 44, 45, 46, 47  
-  };
   static uint8_t __attribute__ ((unused)) the_sockets[] = 
   {
     0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
@@ -252,7 +244,7 @@ extern "C" {
   //debugging functions
 #ifdef DEBUG
 #  define DPRINT(args...) fprintf(stderr,args);
-#  define DDPRINT(fmt, args...) printf("%s:%s:%d: "fmt, __FILE__, __FUNCTION__, __LINE__, args)
+#  define DDPRINT(fmt, args...) printf("%s:%s:%d: " fmt, __FILE__, __FUNCTION__, __LINE__, args)
 #else
 #  define DPRINT(...)
 #  define DDPRINT(fmt, ...)
@@ -346,7 +338,7 @@ extern "C" {
 
   //getticks needs to have a correction because the call itself takes a
   //significant number of cycles and skewes the measurement
-  extern inline ticks getticks_correction_calc();
+  extern ticks getticks_correction_calc();
 
   static inline ticks get_noop_duration() {
 #define NOOP_CALC_REPS 1000000

### 2014-03-18 paper/paper.tex
Paper edits

@@
 \definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
+
+\pdfstringdefDisableCommands{\def\citeauthor#1{#1}}
 
@@
 
-Concurrent priority queues have been the subject of research since the 1980s.
+Concurrent priority queues have been the subject of research since the 1980s
+\cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
+luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
 While early efforts have focused mostly on parallelizing Heap structures
@@
 \cite{pugh1990skip} seem to show more potential \cite{shavit2000skiplist,
-sundell2003fast,herlihy2012art,linden2013skiplist}.
+sundell2003fast,herlihy2012art,linden2013skiplist}. Current research has also examined
+relaxed data structures \cite{wimmer2013data,alistarhspraylist} which trade
+the strictness of provided guarantees against improved scalability.
 
@@
 
-In the following, we examine the evolution of concurrent priority queues. We begin
-with first steps which used fine-grained locking to avoid the bottleneck of a single
-global lock. Initial SkipList-based structures offered better disjoint-access
-parallelism, and finally current state of the art lock-free priority queues minimize
-contention in \lstinline|DeleteMin|.
+In the following, we examine the evolution of concurrent priority queues suitable for general use,
+i.e. unbounded range queues based on widely available atomic primitives such as \ac{CAS}.
+Section \ref{sec:concepts} outlines basic concepts and definitions. In Section \ref{sec:hunt},
+we cover the \citeauthor{hunt1996efficient} queue as a representative of early heap-based queues using
+fine-grained locking to avoid the bottleneck of a single global lock. Lock-free SkipList-based
+structures offering better disjoint-access parallelism are discussed in Section \ref{sec:lockfree},
+while Section \ref{sec:relaxed} introduces the concept of relaxed data structures and presents two
+such priority queues. Finally, related work is presented in Section \ref{sec:related}, and the paper is
+concluded in Section \ref{sec:conclusion}.
 
-% TODO: Mention what we don't cover: non-general PQs, unavailable primitives, ...
-% TODO: In section x, we ... Again, this partially duplicates the last paragraph.
-
-\section{Concepts and Definitions}
+\section{Concepts and Definitions} \label{sec:concepts}
 
@@
 \emph{Wait-freedom} further guarantees that every process finishes each operation in a bounded number of steps. 
-In practise, wait-freedom often introduces an unacceptable overhead; lock-freedom
+In practice, wait-freedom often introduces an unacceptable overhead; lock-freedom
 however has proven to be both efficient and to scale well to large numbers of processes.
+Recently, \citeauthor{kogan2012methodology} have also developed a methodology for implementing efficient
+wait-free data structures \cite{kogan2012methodology}.
 
@@
 An area in memory accessed frequently by a large number of processes is said to be \emph{contended}.
-Contention is an limiting factor regarding scalability: concurrent reads and writes to the same 
+Contention is a limiting factor regarding scalability: concurrent reads and writes to the same
 location must be serialized by the cache coherence protocol, and only a single concurrent \ac{CAS} % TODO: Short section on primitives.
@@
 SkipLists \cite{pugh1990skip} have become increasingly popular in concurrent data structures because 
-they are both simple and exhibit excellent disjoint-access parallelism. In contrast to
+they are both simple to implement, and exhibit excellent disjoint-access parallelism. In contrast to
 search trees and Heaps, reorganization is not necessary, since SkipLists rely on
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{lstlisting}
@@
 
+In the remaining paper, we will discuss implementations of several concurrent priority queue implementations.
+Early designs have mostly been based on search trees \cite{boyar1994chromatic,johnson1991highly} and
+heaps \cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
+luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
 We chose the priority queue by \citeauthor{hunt1996efficient} \cite{hunt1996efficient}
@@
 every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
-Finally, in contrast to later SkipList-based designs, the capacity of Hunt Heap
+Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heap
 is fixed upon creation.
@@
 the size of the Heap reaches a certain threshold such that concurrency
-can be properly exploited (instead of being dominated by global locking overhead).
+can be properly exploited instead of being dominated by global locking overhead.
 
@@
 SkipLists are both conceptual simple as well as simple to implement; they also exhibit
-excellent disjoint-access properties, and do not require rebalancing due to their
-reliance on randomization for expected $O(\log n)$ time insertion and deletion bounds.
+excellent disjoint-access parallelism properties, and do not require rebalancing due to their
+reliance on randomization.
 
-A state of the art lock-free SkipList implementation based on \ac{CAS}
+A state of the art lock-free SkipList implementation based on the \ac{CAS} instruction
 by \citeauthor{fraser2004practical} \cite{fraser2004practical} is freely available\footnote{
@@
 are required. As an additional requirement, these memory management schemes must
-be both scalable and lock-free to avoid limiting the SkipList itself.
-\citeauthor{fraser2004practical} in particular employs lock-free epoch-based garbage-collection.
+themselves be both scalable and lock-free to avoid limiting the SkipList itself.
+\citeauthor{fraser2004practical} in particular employs lock-free epoch-based garbage-collection,
+which frees a memory segment only once all threads that could have seen a pointer to it have
+exited the data structure.
 
@@
 \citeauthor{shavit2000skiplist} were the first to propose the use of SkipLists
-for priority queues \cite{linden2013skiplist}. Their initial implementation
+for priority queues \cite{linden2013skiplist}. Their initial locking implementation
 \cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
-SkipList \cite{pugh1998concurrent} which uses one lock per node per level.
+SkipList \cite{pugh1998concurrent}, which uses one lock per node per level.
 
@@
 Therefore, both insertions and deletions can be split into steps --- insertions
-proceed bottom-up while deletions proceed top-down, holding locks only for the current level.
+proceed bottom-up while deletions proceed top-down. Locks are held only for the current level
+which helps to reduce contention between threads.
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{lstlisting}
@@
 
-Likewise, deletions are split into a logical phase, atomically setting the \lstinline|node.deleted|
-flag; and a physical phase performs the actual pointer manipulations and can be seen as a simple
+Likewise, deletions are split into a logical phase (atomically setting the \lstinline|node.deleted|
+flag) and a physical phase which performs the actual pointer manipulations and can be seen as a simple
 call to the underlying SkipList's \lstinline|sl_delete| function.
@@
 slow thread A is in the middle of a \lstinline|DeleteMin| call. Within this context, we refer to 
-the node with key $i$ as ``node $i$'', or simply $i$. Several \lstinline|CAS|
+the node with key $i$ as node $i$, or simply $i$. Several \lstinline|CAS|
 operations have failed, and A is currently at node $j$.
@@
 \citeauthor{shavit2000skiplist} counteract this by introducing a \lstinline|timestamp| for each
-node which is set upon successful insertion. Each \lstinline|DeleteMin| operation 
+node which is set upon successful insertion. In this variant, each \lstinline|DeleteMin| operation
 simply ignores all nodes it sees that have not been fully inserted at the time it was called.
@@
 
-\citeauthor{herlihy2012art} \cite{herlihy2012art} later described and implemented a variant of this idea 
-in Java that is lock-free and quiescently consistent. While mostly identical, notable differences are
+\citeauthor{herlihy2012art} \cite{herlihy2012art} recently described and implemented a lock-free,
+quiescently consistent version of this idea in Java. While mostly identical, notable differences are
 that a) the new variant is based on a lock-free skiplist, b) the timestamping mechanism was not
@@
 In contrast to other structures covered in this paper, this priority queue is restricted to
-contain items with pairwise distinct priorities.
+contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
+in the list simply performs an update of the associated value.
 A real-time version is also provided which will we will not discuss further (interested readers are
@@
 exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
-are unused and reuse pointers as deletion marks. A set least significant bit signifies on a pointer signifies
-that the current node is about to be deleted. We will use the notation 
-\lstinline|CAS(node.level[i], (ptr, false), (ptr, true))| to signify \ac{CAS} operations on the
-combined pointer \& flag fields.
+are unused and reuse these as deletion marks. A set least significant bit signifies on a pointer signifies
+that the current node is about to be deleted.
 Reuse of \lstinline|node.level[i]| pointers
@@
 A rigorous linearizability proof is provided in the original paper \cite{sundell2003fast} which shows
-linearizability points for all possible outcomes of both \lstinline|Insert| and \lstinline|DeleteMin|
-operations.
+linearization points for all possible outcomes of all operations.
 
@@
 in \citeyear{linden2013skiplist} \cite{linden2013skiplist}. They present a linearizable, lock-free concurrent priority
-queue which achieves a speed-up of $30-80\%$ compared to other SkipList-based concurrent priority queues by
+queue which achieves a speed-up of $30-80\%$ compared to other SkipList-based priority queues by
 minimizing the number of \ac{CAS} operations within most \lstinline|DeleteMin| operations.
@@
 It is called strict when \lstinline|DeleteMin| is guaranteed to return the minimal element currently within the queue
-(as opposed to relaxed data structures which are discussed further in the next section).
+(in contrast to relaxed data structures which are discussed further in the next section).
 All such priority queues share an inherent bottleneck, since all threads calling \lstinline|DeleteMin| compete
@@
 
-Once the prefix of logically deleted nodes reaches a specified length {\lstset{breaklines,breakatwhitespace} \lstinline|BoundOffset|},
+Once the prefix of logically deleted nodes reaches a specified length (represented by \lstinline|BoundOffset|),
 the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
@@
 
-\section{Relaxed Priority Queues}
+\section{Relaxed Priority Queues} \label{sec:relaxed}
 
@@
 
-Whenever an item is added, it is first added to both \lstinline|local_list| and \lstinline|prio_queue|
+Whenever an item is added, it is first added locally to both \lstinline|local_list| and \lstinline|prio_queue|
 and \lstinline|remaining_k| is decremented. If \lstinline|remaining_k| reaches zero, then the local
@@
 
-\section{Related Work}
+\section{Related Work} \label{sec:related}
 
@@
 
-\section{Conclusion}
+\section{Conclusion} \label{sec:conclusion}
 

### 2014-03-18 paper/paper.tex
Conclusion

@@
 
+Priority queues are one of the most important abstract data structures in computer science,
+and much effort has been put into parallelizing them efficiently. In this paper,
+we have outlined the evolution of concurrent priority queues from initial heap-based designs,
+through a period of increasingly efficient SkipList queues, to current research into relaxed data structures.
+
+The switch from heaps to SkipLists as the backing data structure highlights how a simple change in
+direction can help revitalize an entire field of research. SkipList-based priority queues are the current
+state of the art in strict shared-memory concurrent priority queues: they provide strong guarantees
+and scale well to up to the ten's of threads in practice. Important limiting factors are contention
+at the front of the list and the large number of \ac{CAS} failures.
+The \citeauthor{linden2013skiplist} queue is designed
+to minimize the latter factor; but the former is inherent to all strict priority queues, which could mean
+that the peak performance in such structures has been reached.
+
+Recently invented relaxed priority queues do not exhibit the inherent bottleneck at the front of the list,
+as they do not necessarily return the minimal element within the queue and are able to spread
+\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeable
+better and to larger thread counts than strict designs. Further research is necessary in order to fully
+explore the possibilities provided by relaxed data structures.
+
 \begin{comment}

### 2014-03-18 paper/paper.tex
Abstract

@@
 \begin{abstract}
-On the 24th of February, 1815, the watch-tower of Notre-Dame de la Garde
-signalled the arrival of the three-master Pharaon, from Smyrna, Trieste, 
-and Naples.
+Priority queues are abstract data structures which store a set of (priority, item) pairs
+and allow efficient access to the item with minimal priority. Such queues are an important
+element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
+path algorithm) and operating system (priority schedulers).
 
-The usual crowd of curious spectators immediately filled the quay of Fort 
-Saint-Jean, for at Marseilles the arrival of a ship is always a great event, 
-especially when that ship, as was the case with the Pharaon, has been built, 
-rigged, and laden in the dockyard of old Phocaea and belongs to a shipowner 
-of their own town.
+The recent trend towards multiprocessor computing requires new implementations of basic
+data structures which are able to be used concurrently and ideally scale well to a large number
+of threads. In particular, lock-free structures promise superior scalability by avoiding
+the use of blocking synchronization primitives.
 
-Meanwhile the vessel drew on, and was approaching the harbour under topsails, 
-jib, and foresail, but so slowly and with such an air of melancholy that the 
-spectators, always ready to sense misfortune, began to ask one another what 
-ill-luck had overtaken those on board. However, those experienced in navigation 
-soon saw that if there had been any ill-luck, the ship had not been the 
-sufferer, for she advanced in perfect condition and under skilful handling; 
-the anchor was ready to be dropped, the bowsprit shrouds loose. 
-Beside the pilot, who was steering the Pharaon through the narrow entrance to 
-the port, there stood a young man, quick of gesture and keen of eye, who 
-watched every movement of the ship while repeating each of the pilot's orders.
+Concurrent priority queues have been extensively researched over the past decades.
+In this paper, we discuss three major ideas within the field: fine-grained locking
+employs multiple locks to avoid a single bottleneck within the queue; SkipLists
+are search structures which use randomization and therefore do not require elaborate reorganization
+schemes; and relaxed data structures trade semantic guarantees for improved scalability.
 \end{abstract}

### 2014-03-20 paper/paper.tex
DeleteMin removes lowest priority item

@@
 places an item into the queue together with its priority, while \lstinline|DeleteMin|
-removes and returns the highest priority item. Both of these operations are expected to have
+removes and returns the lowest priority item. Both of these operations are expected to have
 a complexity of at most $O(\log n)$. Priority queues are used in a large variety
@@
 Priority queues are an abstract data structure allowing insertion of items with a given priority
-and removal of the highest-priority item in logarithmic time. Search trees and Heaps (which are 
+and removal of the lowest-priority item in logarithmic time. Search trees and Heaps (which are
 flattened representations of complete trees such that each node's value is at least as large as

### 2014-03-20 paper/paper.tex
Paper edits

@@
 element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
-path algorithm) and operating system (priority schedulers).
+path algorithm) and operating system (i.e. priority schedulers).
 
 The recent trend towards multiprocessor computing requires new implementations of basic
-data structures which are able to be used concurrently and ideally scale well to a large number
+data structures which are able to be used concurrently and scale well to a large number
 of threads. In particular, lock-free structures promise superior scalability by avoiding
@@
 relaxed data structures \cite{wimmer2013data,alistarhspraylist} which trade
-the strictness of provided guarantees against improved scalability.
+strictness of provided guarantees against improved scalability.
 
@@
 
-In the following, we examine the evolution of concurrent priority queues suitable for general use,
+In the following, we investigate the evolution of concurrent priority queues suitable for general use,
 i.e. unbounded range queues based on widely available atomic primitives such as \ac{CAS}.
@@
 There have been many publications on more specialized variants of concurrent priority queues
-which we did not cover in this paper.
+which were not covered in this paper.
 
@@
 \citeauthor{liu2012lock} present a lock-free, array-based priority queue in \cite{liu2012lock}
-using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the standard
+using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the
 \lstinline|Insert| and \lstinline|DeleteMin| operations, it also provides an \lstinline|ExtractMany|
@@
 \citeauthor{shavit1999scalable} present two such data structures in \cite{shavit1999scalable} using
-combining funnels and bins of items.
+combining funnels and bins of items. Experimental results have shown strong scalability until at least
+256 threads.
 
@@
 \citeauthor{kirsch2012fast} invented an efficient k-FIFO queue \cite{kirsch2012fast},
-% TODO: More cites.
+The trend towards relaxed data structures in general is examined further in
+\cite{shavit2011data,kirsch2012incorrect}.
 
@@
 state of the art in strict shared-memory concurrent priority queues: they provide strong guarantees
-and scale well to up to the ten's of threads in practice. Important limiting factors are contention
+and scale well to up to the tens of threads in practice. Important limiting factors are contention
 at the front of the list and the large number of \ac{CAS} failures.
 The \citeauthor{linden2013skiplist} queue is designed
-to minimize the latter factor; but the former is inherent to all strict priority queues, which could mean
+to minimize the latter; but the former is inherent to all strict priority queues, which could mean
 that the peak performance in such structures has been reached.

### 2014-03-20 src/pqbench.cpp
Switch back to initial queue size of 2^15

@@ -16,7 +16,7 @@
 #define DEFAULT_SECS     (10)
 #define DEFAULT_NTHREADS (1)
 #define DEFAULT_OFFSET   (128)
-#define DEFAULT_SIZE     (1 << 20)
+#define DEFAULT_SIZE     (1 << 15)
 #define DEFAULT_VERBOSE  (false)
 
 static std::atomic<bool> loop;

### 2014-03-20 presentation/presentation.tex
An initially empty presentation.tex

--- 2014-03-20 paper/paper.tex
@@
-\documentclass[a4paper,10pt]{article}
+\documentclass{beamer}
 
@@
 \usepackage[backend=bibtex]{biblatex}
-\usepackage[usenames,dvipsnames]{color}
-\usepackage{comment}
-\usepackage[pdfborder={0 0 0}]{hyperref}
+\usepackage{default}
 \usepackage[utf8]{inputenc}
-\usepackage{listings}
+\usepackage{lmodern}
 
@@
 
-\definecolor{Gray}{gray}{0.5}
-\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
+\setbeamertemplate{bibliography item}{}
 
-\pdfstringdefDisableCommands{\def\citeauthor#1{#1}}
-
-\lstset{
-    language=C++,
-    basicstyle=\ttfamily,
-    keywordstyle=\color{OliveGreen},
-    commentstyle=\color{Gray},
-    captionpos=b,
-    breaklines=true,
-    breakatwhitespace=false,
-    showspaces=false,
-    showtabs=false,
-    numbers=left,
-}
-
-\title{State of the Art in Concurrent Priority Queues \\
-       Student Paper for Seminar in Algorithms 2013W \\
-       Technical University of Vienna}
+\title{Concurrent Priority Queues}
+\subtitle{Seminar in Algorithms, 2013W}
 \author{Jakob Gruber, 0203440}
+\date{\today}
 
@@
 
-\begin{comment}
-Abstract formula
-----------------
+\begin{frame}{Hello World}
+Hello
+\end{frame}
 
-1) big picture problem or topic widely debated in your field.
-2) gap in the literature on this topic.
-3) your project filling the gap.
-4) the specific material that you examine in the paper.
-5) your original argument.
-6) a strong concluding sentence.
-
-Introduction formula
---------------------
-
-A good paper introduction is fairly formulaic. If you follow a simple set of
-rules, you can write a very good introduction. The following outline can be
-varied. For example, you can use two paragraphs instead of one, or you can
-place more emphasis on one aspect of the intro than another. But in all cases,
-all of the points below need to be covered in an introduction, and in most
-papers, you don't need to cover anything more in an introduction.
-
-Paragraph 1: Motivation. At a high level, what is the problem area you are
-working in and why is it important? It is important to set the larger context
-here. Why is the problem of interest and importance to the larger community?
-
-Paragraph 2: What is the specific problem considered in this paper? This
-paragraph narrows down the topic area of the paper. In the first paragraph you
-have established general context and importance. Here you establish specific
-context and background.
-
-Paragraph 3: "In this paper, we show that ...". This is the key paragraph in
-the intro - you summarize, in one paragraph, what are the main contributions of
-your paper given the context you have established in paragraphs 1 and 2. What
-is the general approach taken? Why are the specific results significant? This
-paragraph must be really really good. If you can't "sell" your work at a high
-level in a paragraph in the intro, then you are in trouble. As a reader or
-reviewer, this is the paragraph that I always look for, and read very
-carefully.
-
-You should think about how to structure this one or two paragraph summary of
-what your paper is all about. If there are two or three main results, then you
-might consider itemizing them with bullets or in test (e.g., "First, ..."). If
-the results fall broadly into two categories, you can bring out that
-distinction here. For example, "Our results are both theoretical and applied in
-nature. (two sentences follow, one each on theory and application)"
-
-Paragraph 4: At a high level what are the differences in what you are doing,
-and what others have done? Keep this at a high level, you can refer to a future
-section where specific details and differences will be given. But it is
-important for the reader to know at a high level, what is new about this work
-compared to other work in the area.
-
-Paragraph 5: "The remainder of this paper is structured as follows..." Give the
-reader a roadmap for the rest of the paper. Avoid redundant phrasing, "In
-Section 2, In section 3, ... In Section 4, ... " etc.
-
-A few general tips:
-
-Don't spend a lot of time into the introduction telling the reader about what
-you don't do in the paper. Be clear about what you do do, but don't dwell here
-on what you don't do.  Does each paragraph have a theme sentence that sets the
-stage for the entire paragraph? Are the sentences and topics in the paragraph
-all related to each other?  Do all of your tenses match up in a paragraph?
-\end{comment}
-
-\begin{abstract}
-Priority queues are abstract data structures which store a set of (priority, item) pairs
-and allow efficient access to the item with minimal priority. Such queues are an important
-element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
-path algorithm) and operating system (priority schedulers).
-
-The recent trend towards multiprocessor computing requires new implementations of basic
-data structures which are able to be used concurrently and ideally scale well to a large number
-of threads. In particular, lock-free structures promise superior scalability by avoiding
-the use of blocking synchronization primitives.
-
-Concurrent priority queues have been extensively researched over the past decades.
-In this paper, we discuss three major ideas within the field: fine-grained locking
-employs multiple locks to avoid a single bottleneck within the queue; SkipLists
-are search structures which use randomization and therefore do not require elaborate reorganization
-schemes; and relaxed data structures trade semantic guarantees for improved scalability.
-\end{abstract}
-
-\section{Introduction}
-
-% Move to many cores, why do we need concurrent algorithms/data structures?
-
-In the past decade, advancements in computer performance have been made mostly
-through an increasing number of processors instead of higher clock speeds.
-This development necessitates new approaches to data structures and algorithms
-that take advantage of concurrent execution on multiple threads and processors.
-
-% The priority queue - semantics and usage summary.
-
-This paper focuses on the priority queue data structure, consisting of two operations
-traditionally called \lstinline|Insert| and \lstinline|DeleteMin|. \lstinline|Insert|
-places an item into the queue together with its priority, while \lstinline|DeleteMin|
-removes and returns the lowest priority item. Both of these operations are expected to have
-a complexity of at most $O(\log n)$. Priority queues are used in a large variety
-of situations such as shortest path algorithms and scheduling. % See liu2012lock for more.
-
-% Concurrent priority queues - heaps, trees, skiplists.
-
-Concurrent priority queues have been the subject of research since the 1980s
-\cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
-luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
-While early efforts have focused mostly on parallelizing Heap structures
-\cite{hunt1996efficient}, % TODO: Many more, see \cite{shavit2000skiplist}
-more recently priority queues based on \citeauthor{pugh1990skip}'s SkipLists
-\cite{pugh1990skip} seem to show more potential \cite{shavit2000skiplist,
-sundell2003fast,herlihy2012art,linden2013skiplist}. Current research has also examined
-relaxed data structures \cite{wimmer2013data,alistarhspraylist} which trade
-the strictness of provided guarantees against improved scalability.
-
-% Our plan for the paper - somehow this kind of duplicates the last paragraph though.
-
-In the following, we examine the evolution of concurrent priority queues suitable for general use,
-i.e. unbounded range queues based on widely available atomic primitives such as \ac{CAS}.
-Section \ref{sec:concepts} outlines basic concepts and definitions. In Section \ref{sec:hunt},
-we cover the \citeauthor{hunt1996efficient} queue as a representative of early heap-based queues using
-fine-grained locking to avoid the bottleneck of a single global lock. Lock-free SkipList-based
-structures offering better disjoint-access parallelism are discussed in Section \ref{sec:lockfree},
-while Section \ref{sec:relaxed} introduces the concept of relaxed data structures and presents two
-such priority queues. Finally, related work is presented in Section \ref{sec:related}, and the paper is
-concluded in Section \ref{sec:conclusion}.
-
-\section{Concepts and Definitions} \label{sec:concepts}
-
-% Blocking, lock-free and wait-free data structures.
-
-Concurrent data structures are intended to be accessed simultaneously by several processes
-at once. \emph{Lock-based} structures ensure that only a limited number of processes may enter
-a critical section at once. \emph{Lock-free} data structures eschew the use of locks, and guarantee
-that at least a single process makes progress at all times. Since lock-free structures are
-non-blocking, they are not susceptible to priority inversion, deadlock, and livelock.
-\emph{Wait-freedom} further guarantees that every process finishes each operation in a bounded number of steps. 
-In practice, wait-freedom often introduces an unacceptable overhead; lock-freedom
-however has proven to be both efficient and to scale well to large numbers of processes.
-Recently, \citeauthor{kogan2012methodology} have also developed a methodology for implementing efficient
-wait-free data structures \cite{kogan2012methodology}.
-
-% Linearizability, sequential and quiescent consistency.
-
-There are several different criteria which allows reasoning about the correctness of concurrent
-data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
-effect at a single instant in time at so-called linearization points. Linearizability is composable ---
-any data structure composed of linearizable objects is also linearizable.
-\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
-executed in parallel to be equivalent to the result of some sequential ordering of the same
-operations.
-\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
-guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
-after a period of quiescence, i.e. an interval without active operations has passed.
-
-% Disjoint-access parallelism.
-
-An area in memory accessed frequently by a large number of processes is said to be \emph{contended}.
-Contention is a limiting factor regarding scalability: concurrent reads and writes to the same
-location must be serialized by the cache coherence protocol, and only a single concurrent \ac{CAS} % TODO: Short section on primitives.
-can succeed while all others must retry. \emph{Disjoint-access parallelism} is the concept of
-spreading such accesses in order to reduce contention as much as possible.
-
-% Priority queues.
-
-Priority queues are an abstract data structure allowing insertion of items with a given priority
-and removal of the lowest-priority item in logarithmic time. Search trees and Heaps (which are
-flattened representations of complete trees such that each node's value is at least as large as
-those of both children) are concrete data structures which are usually used to implement sequential
-priority queues. However, both require fairly elaborate reorganization after \lstinline|DeleteMin|
-and/or \lstinline|Insert| operations, which are especially challenging to achieve in a concurrent
-environment.
-
-% SkipLists.
-
-SkipLists \cite{pugh1990skip} have become increasingly popular in concurrent data structures because 
-they are both simple to implement, and exhibit excellent disjoint-access parallelism. In contrast to
-search trees and Heaps, reorganization is not necessary, since SkipLists rely on
-randomization for an expected $O(\log n)$ complexity of \lstinline|Insert|, \lstinline|Search|
-and \lstinline|Delete| operations.
-A SkipList may be visualized as a set of linked lists with corresponding levels. The linked
-list at level 0 is the sorted sequence of all objects in the SkipList,
-and higher levels provide ``shortcuts''
-into the SkipList such that a list of level $i + 1$ contains a subset of the objects in level
-$i$. A SkipList node $n$ is said to be of level $i$ if it is in all lists of levels $[0, i]$
-and in none of levels $[i + 1, \infty]$. Upon insertion, the new node's level is assigned at random
-according to a geometric distribution; deletion simply removes the node.
-
-% TODO: Maybe integrate into description - otherwise remove. Do we need further detail?
-
-\begin{figure}[ht]
-\begin{lstlisting}
-struct slist_t {
-  size_t max_level;
-  node_t head[max_level];
-};
-
-struct node_t {
-  key_t key;
-  value_t value;
-  size_t level;
-  node_t *next[level];
-};
-
-/* All operations are expected O(log n) time. */
-void slist_insert(slist_t *l, key_t k, value_t v);
-bool slist_delete(slist_t *l, key_t k, value_t &v);
-\end{lstlisting}
-\caption{Basic SkipList structure and operations.}
-\label{fig:basicsl}
-\end{figure}
-
-\section{Fine-grained Locking Heaps} \label{sec:hunt}
-
-In the remaining paper, we will discuss implementations of several concurrent priority queue implementations.
-Early designs have mostly been based on search trees \cite{boyar1994chromatic,johnson1991highly} and
-heaps \cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
-luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
-We chose the priority queue by \citeauthor{hunt1996efficient} \cite{hunt1996efficient}
-as a representative of early concurrent priority queues since it has been proven to
-perform well \cite{shavit2000skiplist} in comparison to other efforts of the time such as \cite{nageshwara1988concurrent,ayani1990lr,yan1998lock}. % Many others \cite{shavit2000skiplist}.
-It is based on a Heap
-structure and attempts to minimize lock contention between threads by a) adding per-node
-locks, b) spreading subsequent insertions through a bit-reversal technique, % Elaborate on these.
-and c) letting insertions traverse bottom-up in order to minimize conflicts with
-top-down deletions.
-
-However, significant limitations to scalability remain. A global lock is required
-to protect accesses to a variable storing the Heap's size which all operations
-must obtain for a short time. Disjoint-access through bit-reversal breaks down
-once a certain amount of traffic is reached, since only subsequent insertions
-are guaranteed to take disjoint paths towards the root node. Note also that
-the root node is a severe serial bottleneck, since it is potentially part of
-every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
-Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heap
-is fixed upon creation.
-
-Benchmarking results in the literature have been mixed; a sequential priority
-queue protected by a single global lock outperforms the \citeauthor{hunt1996efficient}
-Heap in most cases \cite{hunt1996efficient,sundell2003fast}. Speed-up only occurs once
-the size of the Heap reaches a certain threshold such that concurrency
-can be properly exploited instead of being dominated by global locking overhead.
-
-\section{Lock-free Priority Queues} \label{sec:lockfree}
-
-Traditional data structures such as the Heap have fallen out of favor;
-instead, SkipLists \cite{pugh1990skip,pugh1998concurrent} have become the focus
-of modern concurrent priority queue research
-\cite{shavit2000skiplist,sundell2003fast,herlihy2012art,linden2013skiplist,alistarhspraylist}.
-SkipLists are both conceptual simple as well as simple to implement; they also exhibit
-excellent disjoint-access parallelism properties, and do not require rebalancing due to their
-reliance on randomization.
-
-A state of the art lock-free SkipList implementation based on the \ac{CAS} instruction
-by \citeauthor{fraser2004practical} \cite{fraser2004practical} is freely available\footnote{
-\url{http://www.cl.cam.ac.uk/research/srg/netos/lock-free/}} under a BSD license.
-\citeauthor{fraser2004practical} exploits unused pointer bits to mark nodes as logically
-deleted, with physical deletion following as a second step.
-
-SkipLists are dynamic data structures in the sense that they grow and shrink
-at runtime. In consequence, careful handling of memory accesses and (de)allocations
-are required. As an additional requirement, these memory management schemes must
-themselves be both scalable and lock-free to avoid limiting the SkipList itself.
-\citeauthor{fraser2004practical} in particular employs lock-free epoch-based garbage-collection,
-which frees a memory segment only once all threads that could have seen a pointer to it have
-exited the data structure.
-
-\subsection{\citeauthor{shavit2000skiplist}} \label{sec:shavit}
-
-\citeauthor{shavit2000skiplist} were the first to propose the use of SkipLists
-for priority queues \cite{linden2013skiplist}. Their initial locking implementation
-\cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
-SkipList \cite{pugh1998concurrent}, which uses one lock per node per level.
-
-% TODO: This counts for all skiplist impls.
-
-A crucial observation is that nodes nodes which are only partially connected
-do not affect correctness of the data structure. As soon as the first level (i.e. \lstinline|node.level[0]|)
-has been successfully connected, a node is considered to be in the SkipList.
-Therefore, both insertions and deletions can be split into steps --- insertions
-proceed bottom-up while deletions proceed top-down. Locks are held only for the current level
-which helps to reduce contention between threads.
-
-\begin{figure}[ht]
-\begin{lstlisting}
-struct node_t {
-  [...] /**< Standard node members as above. */
-  atomic<bool> deleted; /**< Initially false. */
-  time_t timestamp;
-  lock_t locks[level + 1];
-};
-\end{lstlisting}
-\caption{\citeauthor{shavit2000skiplist} structure.}
-\label{fig:shavitsl}
-\end{figure}
-
-Likewise, deletions are split into a logical phase (atomically setting the \lstinline|node.deleted|
-flag) and a physical phase which performs the actual pointer manipulations and can be seen as a simple
-call to the underlying SkipList's \lstinline|sl_delete| function.
-
-A \lstinline|DeleteMin| call starts the list head, and attempts to atomically set
-the deletion flag using a \lstinline|CAS(node.deleted, false, true)| call (or equivalent constructs).
-If it succeeds, the current node is physically deleted and returned to the caller. Otherwise,
-\lstinline|node.next[0]| is set as the new current node and the procedure is repeated.
-If the end of the list is reached, \lstinline|DeleteMin| returns false to indicate an empty list.
-
-% Logical/physical deletion separation allows concurrent physical deletions.
-
-Note that so far, this implementation is not linearizable: consider the case in which a
-slow thread A is in the middle of a \lstinline|DeleteMin| call. Within this context, we refer to 
-the node with key $i$ as node $i$, or simply $i$. Several \lstinline|CAS|
-operations have failed, and A is currently at node $j$.
-A fast thread B then first inserts a node $i$, followed by a node $k$ such that
-$i < j < k$, i.e. the former and latter nodes are inserted, respectively, before and after
-thread A's current node. Assuming further that all nodes between $j$ and $k$ have already
-been deleted, then thread A will return node $k$. This execution is not linearizable; however,
-it is quiescently consistent since operations can be reordered to correspond to some sequential
-execution at quiescent periods.
-
-\citeauthor{shavit2000skiplist} counteract this by introducing a \lstinline|timestamp| for each
-node which is set upon successful insertion. In this variant, each \lstinline|DeleteMin| operation
-simply ignores all nodes it sees that have not been fully inserted at the time it was called.
-
-% TODO: Linearizability points.
-
-Explicit memory management is required to avoid dereferencing pointers to freed memory areas
-by other threads after physical deletion. This implementation uses a dedicated garbage collector
-thread in combination with a timestamping mechanism which frees \lstinline|node|'s memory only
-when all threads that might have seen a pointer to \lstinline|node| have exited the data structure.
-
-\citeauthor{herlihy2012art} \cite{herlihy2012art} recently described and implemented a lock-free,
-quiescently consistent version of this idea in Java. While mostly identical, notable differences are
-that a) the new variant is based on a lock-free skiplist, b) the timestamping mechanism was not
-employed and thus linearizability was lost, and c) explicit memory management is not required
-because the Java virtual machine provides a garbage collector.
-
-\subsection{\citeauthor{sundell2003fast}} \label{sec:sundell}
-
-\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
-\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
-and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
-In contrast to other structures covered in this paper, this priority queue is restricted to
-contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
-in the list simply performs an update of the associated value.
-A real-time version is also provided which will we will not discuss further (interested readers are
-referred to \cite{sundell2003fast}).
-
-\begin{figure}[ht]
-\begin{lstlisting}
-struct node_t {
-  [...] /**< Standard node members as above. */
-  size_t valid_level;
-  node_t *prev;
-};
-\end{lstlisting}
-\caption{\citeauthor{sundell2003fast} structure.}
-\label{fig:sundellsl}
-\end{figure}
-
-The structure of each node is basically identical to Figure \ref{fig:basicsl}. However, \citeauthor{sundell2003fast}
-exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
-are unused and reuse these as deletion marks. A set least significant bit signifies on a pointer signifies
-that the current node is about to be deleted.
-Reuse of \lstinline|node.level[i]| pointers
-prevents situations in which a new node is inserted while its predecessor is being removed,
-effectively deleting both nodes from the list. Likewise, the reuse of the \lstinline|node.value|
-pointer ensures that updates of pointer values (which occur when a node with the inserted priority already exists)
-handle concurrent node removals correctly.
-
-As in the \citeauthor{shavit2000skiplist} priority queue, insertions proceed bottom-up while
-deletions proceed top-down --- on the one hand, the choice of opposite directions reduces collisions
-between concurrent insert and delete operations, while on the other hand removing nodes from top levels first
-allows many other concurrent operations to simply skip these nodes, further improving performance.
-\lstinline|node.valid_level| is updated during inserts to always equal the highest level of the SkipList
-at which pointers in this node have already been set (as opposed to \lstinline|node.level|, which equals
-the final level of the node).
-
-A helping mechanism is employed whenever a node is encountered that has its deletion bit set, which attempts
-to set the deletion bits on all next pointers and then removes the node from the current level. The
-\lstinline|node.prev| pointer is used as a shortcut to the previous node, avoiding a complete retraversal
-of the list.
-
-% TODO: Mention backoff.
-
-This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
-\cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
-\cite{michael1995correction}. It was chosen in particular because this scheme can guarantee validity
-of \lstinline|prev| as well as all \lstinline|next| pointers. Additionally, it does not require a separate
-garbage collector thread.
-
-A rigorous linearizability proof is provided in the original paper \cite{sundell2003fast} which shows
-linearization points for all possible outcomes of all operations.
-
-% TODO: Further details?
-
-Benchmarks performed by \citeauthor{sundell2003fast} show their queue performing noticeably better than both locking 
-queues from Sections \ref{sec:shavit} and \ref{sec:hunt}, and slightly better than a priority queue
-consisting of a SkipList protected by a single global lock.
-
-\subsection{\citeauthor{linden2013skiplist}} \label{sec:linden}
-
-One of the most recent priority queue implementations was published by \citeauthor{linden2013skiplist}
-in \citeyear{linden2013skiplist} \cite{linden2013skiplist}. They present a linearizable, lock-free concurrent priority
-queue which achieves a speed-up of $30-80\%$ compared to other SkipList-based priority queues by
-minimizing the number of \ac{CAS} operations within most \lstinline|DeleteMin| operations.
-
-A priority queue implementation is called deterministic when the algorithm does not contain randomized elements.
-It is called strict when \lstinline|DeleteMin| is guaranteed to return the minimal element currently within the queue
-(in contrast to relaxed data structures which are discussed further in the next section).
-All such priority queues share an inherent bottleneck, since all threads calling \lstinline|DeleteMin| compete
-for the minimal element, causing both contention through concurrent \ac{CAS} operations on the same variable
-as well as serialization effort by the cache coherence protocol for all other processor accessing the same cache
-line.
-
-\begin{figure}[ht]
-\begin{lstlisting}
-struct node_t {
-  [...] /**< Standard node members as above. */
-  atomic<bool> inserting;
-};
-\end{lstlisting}
-\caption{\citeauthor{linden2013skiplist} structure.}
-\label{fig:lindensl}
-\end{figure}
-
-% TODO: Mention inserting flag and memory management!
-
-In this implementation, \lstinline|DeleteMin| operations only perform logical deletion by setting
-the deletion flag with a single \ac{FAO} call; nodes are only deleted physically once a certain
-threshold of logically deleted nodes is reached.
-
-This mechanism requires a new invariant, in that the set of all logically deleted nodes must always
-form a prefix of the SkipList. Recall that in the \citeauthor{sundell2003fast} queue, deletion flags
-for node \lstinline|n| were packed into \lstinline|n.next| pointers, preventing insertion of new
-nodes \emph{after} deleted nodes. This implementation instead places the deletion flag into the
-lowest level \lstinline|next| pointer of the previous node, preventing inserations \emph{before}
-logically deleted nodes.
-
-Once the prefix of logically deleted nodes reaches a specified length (represented by \lstinline|BoundOffset|),
-the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
-deletions by updating \lstinline|slist.head[0]| to point at the last logically deleted node with a
-single \ac{FAO} operation. The remaining \lstinline|slist.head| pointers are then updated, and
-all physically deleted nodes are marked for recycling.
-
-Since at any time, the data structure contains a prefix of logically deleted nodes, all \lstinline|DeleteMin|
-operations must traverse this sequence before reaching a non-deleted node. In general, reads of nonmodified
-memory locations are very cheap; however, benchmarks in \cite{linden2013skiplist} have shown that
-after a certain point, the effort spent in long read sequences significantly outweighs the reduced
-number of \ac{CAS} calls. It is therefore crucial to choose \lstinline|BoundOffset| carefully, with the
-authors recommending a prefix length bound of 128 for 32 threads.
-
-The actual \lstinline|DeleteMin| and \lstinline|Insert| implementations are surprisingly simple.
-Deletions simply traverse the list until the first node for which \lstinline|(ptr, d) = FAO((node.next[0], d), 1)|
-returns a previously unset deletion flag (\lstinline|d = 0|) and then return \lstinline|ptr|.
-Insertions occur bottom-up and follow the basic \citeauthor{fraser2004practical} algorithm \cite{fraser2004practical},
-taking the separation of deletion flags and nodes into account.
-
-The authors provide high level correctness and linearizability proofs as well as a model for the
-SPIN model checker. Performance has been shown to compare favorably to both
-\citeauthor{sundell2003fast} and \citeauthor{shavit2000skiplist} queues, with throughput improved by
-up to $80\%$.
-
-\section{Relaxed Priority Queues} \label{sec:relaxed}
-
-The body of work discussed in previous sections
-creates the impression that the outer limits of strict, deterministic priority queues have been reached.
-In particular, \citeauthor{linden2013skiplist} conclude that scalability is solely limited by \lstinline|DeleteMin|,
-and that less than one modified memory location per thread and operation would have to be read
-in order to achieve improved performance \cite{linden2013skiplist}.
-
-Recently, relaxation of provided guarantees have been investigated as another method of reducing
-contention and improving disjoint-access parallelism.
-For instance, k-FIFO queues \cite{kirsch2012fast} have achieved considerable
-speed-ups compared to strict FIFO queues by allowing {\lstset{breaklines,breakatwhitespace} \lstinline|Dequeue|} to return elements
-up to $k$ positions out of order.
-
-Relaxation has also been applied to concurrent priority queues with some success, and in the following
-sections we discuss two such approaches.
-
-\subsection{\citeauthor{wimmer2013data}} \label{sec:wimmer}
-
-\citeauthor{wimmer2013data} presented the first relaxed, linearizable, and lock-free priority queue
-in \cite{wimmer2013data}. It is integrated as a priority scheduler into their \emph{Pheet} task-scheduling
-system, and an open-source implementation is available\footnote{\url{http://pheet.org}}.
-
-Their paper presents several variations on the common theme of priority queues: a distributed work-stealing
-queue which can give no guarantees as to global ordering since it consists of separate priority queues
-at each thread; a relaxed centralized priority queue in which no more than $k$ items are missed
-by any processor; and a relaxed hybrid data structure which combines both ideas and provides
-a guarantee that no thread misses more than $kP$ items where $P$ is the number of participating threads.
-In this section, we examine only the hybrid variant since it provides both the scalability of work-stealing
-queues and the ordering guarantees of the centralized priority queue.
-
-\begin{figure}[ht]
-\begin{lstlisting}
-struct globals_t {
-  list_of_item_t global_list;
-};
-
-/* Thread-local items. */
-struct locals_t {
-  list_of_item_t local_list;
-  pq_t prio_queue;
-  size_t remaining_k;
-};
-\end{lstlisting}
-\caption{\citeauthor{wimmer2013data} structure.}
-\label{fig:wimmerq}
-\end{figure}
-
-The hybrid queue consists of a list of globally visible items and one local item list as well as a local
-sequential priority queue per thread. The thread-local counter \lstinline|remaining_k| tracks how many more
-items may be added to the local queue until all local items must be made globally visible to avoid
-breaking guarantees.
-
-Whenever an item is added, it is first added locally to both \lstinline|local_list| and \lstinline|prio_queue|
-and \lstinline|remaining_k| is decremented. If \lstinline|remaining_k| reaches zero, then the local
-item list is appended to the global list, and all not yet seen (by this thread) items of the global
-queue are added to the local priority queue.
-
-\lstinline|DeleteMin| simply pops the local priority queue as long is it is non-empty and the popped
-item has already been deleted. When a non-deleted item is popped, it is atomically marked as deleted
-and returned to the caller. If instead we are faced with an empty local queue, we attempt to spy,
-i.e. copy items from another thread's local list.
-
-% TODO: Memory management.
-
-The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
-of Dijkstra's shortest path algorithms. Their model creates a task for each node
-expansion, and therefore comes with a considerable task scheduling overhead. Nonetheless,
-the parallel implementation scales well up to 10 threads, and further limited performance
-gains are made until 40 threads. To date, no direct comparisons to other concurrent
-priority queues have been possible since the data structure is strongly tied to \emph{Pheet},
-but a separate implementation of k-priority queues is planned.
-
-\subsection{SprayList} \label{sec:spraylist}
-
-The SprayList is another recent approach towards a relaxed priority queue by \citeauthor{alistarhspraylist}
-\cite{alistarhspraylist}. Instead of the distributed approach described in the previous section,
-the SprayList is based on a central data structure, and uses a random walk in \lstinline|DeleteMin|
-in order to spread accesses over the $O(P \log^3 P)$ first elements with high probability, where $P$
-is again the number of participating threads.
-
-\citeauthor{fraser2004practical}'s lock-free SkipList \cite{fraser2004practical} again serves as the
-basis for the priority queue implementation. The full source code is available online\footnote{
-\url{https://github.com/jkopinsky/SprayList}}. In the SprayList, \lstinline|Insert| calls are simply
-forwarded to the underlying SkipList.
-
-The \lstinline|DeleteMin| operation however executes a random walk, also called a \emph{spray}, the
-purpose of which is to spread accesses over a certain section of the SkipList uniformly such that
-collisions between multiple concurrent \lstinline|DeleteMin| calls are unlikely. This is achieved by
-starting at some initial height, walking a randomized number of steps, descending a randomized number of levels,
-and repeating this procedure until a node $n$ is reached on the lowest level. If $n$ is not deleted,
-it is logically deleted and returned. Otherwise, a \emph{spray} is either reattempted, or the thread
-becomes a cleaner, traversing the lowest level of the SkipList and physically removing logically deleted
-nodes it comes across. A number of dummy nodes are added to the beginning of the list in order to counteract
-the algorithm's bias against initial items.
-
-The \emph{spray} parameters are chosen such that with high probability, one of the $O(P \log^3 P)$
-first elements is returned, and that each of these elements is chosen roughly uniformly at random.
-The final effect is that accesses to the data structure are spread out, reducing contention and resulting
-in a noticeably lower number of \ac{CAS} failures in comparison to strict priority queues described
-in Section \ref{sec:lockfree}.
-
-The authors do not provide any statement as to the linearizability (or other concurrent correctness
-criteria) of the SprayList, and it is not completely clear how to define it since no
-sequential semantics are given.
-
-Benchmarks show promising results: the SprayList scales well at least up to 80 threads,
-and performs close (within a constant factor) to an implementation using a random remove instead
-of \lstinline|DeleteMin|, which the authors consider as the performance ideal.
-
-% Maybe mention that P log^3 P is a very loose bound.
-
-\section{Performance Results}
-
-In this section, we compare the performance of several different concurrent priority
-queue implementations:
-
-\begin{itemize}
-\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
-      by the C++ standard library, protected by a single global lock.
-\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
-      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
-      However, as the original code was written for the MIPS architecture, we chose to
-      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
-      instead.
-\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
-      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
-      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
-      According to the authors, the commercial Noble library includes a more efficient
-      version of this data structure that can also handle duplicate priority values.
-\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
-      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
-      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
-      SkipList implementation. The aim of this implementation is to minimize contention in
-      calls to \lstinline|DeleteMin|.
-\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
-      SkipList using random walks to spread data accesses
-      incurred by \lstinline|DeleteMin| calls. Code provided by \citeauthor{alistarhspraylist} is
-      available on Github\footnote{\url{https://github.com/jkopinsky/SprayList}}.
-\end{itemize}
-
-In each test run, the examined priority queue was initially filled with $2^{15}$
-elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
-for a total of $10$ seconds, where all \lstinline|Insert|
-operations within this context implicitly choose a key uniformly at random from
-the range of all 32-bit integers. This methodology seems to be the de facto standard for concurrent % TODO: Similar phrasing as linden :(
-priority queue benchmarks
-\cite{alistarhspraylist,linden2013skiplist,shavit2000skiplist,sundell2003fast}.
-Each run was repeated for a total of $10$ times
-and we report on the average throughput.
-
-All benchmarks were compiled with \verb|-O3| using GCC 4.8.2. Evaluations took place
-on an 80-core Intel Xeon machine. The benchmarking code was adapted from \citeauthor{linden2013skiplist}'s
-benchmarking suite and is available at \url{https://github.com/schuay/seminar_in_algorithms}.
-
-% TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
-% Better hardware description - see SprayList section 4, Linden p16.
-% Note: Strong degradation of noble PQ with growing initial size.
-% Standard PQ benchmarking methodology, cite linden + ??.
-% Drop-off when using more than 1 socket, Linden p17.
-
-\begin{comment}
-Sections / rough structure:
-* Basic concepts and definitions. Linearizability, sequ./quiescent consistency,
-  lock-free, wait-free, disjoint-access parallelism (one of the papers has good
-  summaries of these).
-  Maybe atomic primitives such as CAS (but probably not).
-* Priority queue definitions, semantics, usages. Mention inherent
-  non-scalability through DeleteMin().
-* Skiplist, heap definitions, semantics, usages.
-* Follow development of current state of the art from Hunt Heap ~> Shavit ~>
-  Tsigas ~> Linden.
-* Maybe benchmarks on mars.
-
-Additional notes:
-* Sketch development: Heap, Fraser Skiplist (Details, mention why linearizable),
-  Sundell/Tsigas, other Skiplist-based PQs
-* Modern developments: SprayList, Wimmer et al. Relation to other relaxed structures such as k-FIFO.
-\end{comment}
-
-\section{Related Work} \label{sec:related}
-
-There have been many publications on more specialized variants of concurrent priority queues
-which we did not cover in this paper.
-
-% PQ's with non-available primitives
-
-\citeauthor{liu2012lock} present a lock-free, array-based priority queue in \cite{liu2012lock}
-using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the standard
-\lstinline|Insert| and \lstinline|DeleteMin| operations, it also provides an \lstinline|ExtractMany|
-function which returns the $n$ highest priority items in the queue, and a function that returns
-items which ``probably'' have high priority. Other implementations based on non-standard instructions
-can be found in \cite{israeli1993efficient,greenwald1999non}.
-
-% Bounded priority queues
-
-Another possibility is the restriction of priorities to a particular set, i.e. bounded range priorities.
-\citeauthor{shavit1999scalable} present two such data structures in \cite{shavit1999scalable} using
-combining funnels and bins of items.
-
-% PQ's in message passing systems (see SprayList reviews)
-
-Concurrent priority queues have also been studied in other contexts such as distributed memory systems.
-\citeauthor{karp1993randomized} describe a distributed priority queue in which each processor owns
-a local priority queue, inserts are sent to a random priority queue, and \lstinline|DeleteMin| operations
-simply access the local queue \cite{karp1993randomized}.
-\citeauthor{sanders1998randomized} extends this idea to remove the globally best elements instead
-\cite{sanders1998randomized}.
-
-% Other relaxed data structures -> k-FIFO
-
-Finally, the principle of relaxation has also been applied to other data structures.
-\citeauthor{kirsch2012fast} invented an efficient k-FIFO queue \cite{kirsch2012fast},
-% TODO: More cites.
-
-\begin{comment}
-  * hunt: most efficient old-school algorithm
-  * lots of other older references in [4]
-  * israeli, rappoport: wait-free, non-available atomic primitive
-  * lotan, shavit [4]: skiplist, lock-based. first to propose usage of skip-lists [11]
-  * sundell, tsigas [3]: skiplist, lock-free, linearizable, unique priorities
-  * herlihy, shavit in art of multiprocessor programming: based on [4], lock-free.
-  * linden, jonsson [11]: skiplist, lock-free, linearizable, reduced mem contention.
-    first lock-free PQ algorithm [11, 12]
-  * specialized versions such as
-    * bounded priorities: [7] and others
-    * probabilistic extractMin, extractMany: [9]
-\end{comment}
-
-\section{Conclusion} \label{sec:conclusion}
-
-Priority queues are one of the most important abstract data structures in computer science,
-and much effort has been put into parallelizing them efficiently. In this paper,
-we have outlined the evolution of concurrent priority queues from initial heap-based designs,
-through a period of increasingly efficient SkipList queues, to current research into relaxed data structures.
-
-The switch from heaps to SkipLists as the backing data structure highlights how a simple change in
-direction can help revitalize an entire field of research. SkipList-based priority queues are the current
-state of the art in strict shared-memory concurrent priority queues: they provide strong guarantees
-and scale well to up to the ten's of threads in practice. Important limiting factors are contention
-at the front of the list and the large number of \ac{CAS} failures.
-The \citeauthor{linden2013skiplist} queue is designed
-to minimize the latter factor; but the former is inherent to all strict priority queues, which could mean
-that the peak performance in such structures has been reached.
-
-Recently invented relaxed priority queues do not exhibit the inherent bottleneck at the front of the list,
-as they do not necessarily return the minimal element within the queue and are able to spread
-\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeable
-better and to larger thread counts than strict designs. Further research is necessary in order to fully
-explore the possibilities provided by relaxed data structures.
-
-\begin{comment}
-http://leo.stcloudstate.edu/acadwrite/conclude.html
-http://writingcenter.unc.edu/handouts/conclusions/
-http://www.wikihow.com/Write-a-Conclusion-for-a-Research-Paper
-\end{comment}
-
+\begin{frame}[allowframebreaks]{Bibliography}
+\nocite{*}
 \printbibliography
+\end{frame}
 

### 2014-03-21 paper/paper.Rnw
Integrate results plot

--- 2014-03-20 paper/paper.tex
@@
 \usepackage{comment}
+\usepackage{graphicx}
 \usepackage[pdfborder={0 0 0}]{hyperref}
@@
 
+\begin{center}
+<<results = tex, echo = FALSE>>=
+library(Rmisc)
+library(ggplot2)
+library(plyr)
+library(scales)
+
+figfile <- "fig.eps"
+
+# install.packages(c("Rmisc", "ggplot2", "plyr"))
+
+df <- read.csv("../results/results.csv")
+colnames(df) <- c("kernel", "p", "throughput")
+
+df$throughput <- df$throughput/1E6
+
+df2 <- ddply(df, .(kernel, p), summarize, mean = mean(throughput),
+             lower = CI(throughput)[[3]], upper = CI(throughput)[[1]])
+
+# Bar graph
+
+# dodge <- position_dodge(width=0.9)
+# p <- ggplot(df2, aes(x = factor(p), y = mean, fill = factor(kernel))) +
+#             geom_bar(stat = "identity", position = dodge) +
+#             geom_errorbar(aes(ymin = lower, ymax = upper), position = dodge, width = 0.3) +
+#             ylab("throughput [Mops/s]") +
+#             xlab("number of threads")
+
+p <- ggplot(df2, aes(x = p, y = mean, color = kernel, shape = kernel)) +
+            geom_line() +
+            geom_point(size = 4) +
+#            geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
+            ylab("throughput in Mops/s") +
+            xlab("number of threads") +
+# Themes
+            theme_bw() +
+            theme(axis.text = element_text(size = 16),
+                  axis.title = element_text(size = 18),
+                  legend.text = element_text(size = 16),
+                  legend.title = element_text(size = 18),
+                  legend.position = c(1, 1),
+                  legend.justification = c(1, 1),
+                  legend.background = element_rect(fill = alpha("black", 0)))
+
+postscript(file = figfile, paper = "special", width = 10, height = 5, horizontal = FALSE)
+plot(p)
+invisible(dev.off())
+cat("\\includegraphics[width = \\textwidth]{", figfile, "}\n\n", sep = "")
+
+# PDF output
+
+# pdf("abcd.pdf", width=10, height=10)
+# plot(p)
+# dev.off()
+@
+\end{center}
+
 % TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.

### 2014-03-24 bench.py
Properly list all algorithms in bench.py

@@ -41,7 +41,7 @@ def bench(algorithm, ncpus, outfile):
 if __name__ == '__main__':
     parser = OptionParser()
     parser.add_option("-a", "--algorithms", dest = "algorithms", default = ",".join(ALGORITHMS),
-            help = "Comma-separated list of ['heap', 'noble', 'linden']")
+            help = "Comma-separated list of %s" % ALGORITHMS)
     parser.add_option("-n", "--ncpus", dest = "ncpus", default = ",".join(map(str, NCPUS)),
             help = "Comma-separated list of cpu counts")
     parser.add_option("-o", "--outfile", dest = "outfile", default = '/dev/null',

### 2014-03-24 presentation/presentation.tex
Add new sections
Related work
PQ introduction

@@
-\documentclass{beamer}
+\documentclass[usenames,dvipsnames]{beamer}
 
@@
 \usepackage[utf8]{inputenc}
+\usepackage{listings}
 \usepackage{lmodern}
@@
 \setbeamertemplate{bibliography item}{}
+
+\lstset{
+    language=C++,
+    basicstyle=\ttfamily,
+    keywordstyle=\color{OliveGreen},
+    commentstyle=\color{Gray},
+    captionpos=b,
+    breaklines=true,
+    breakatwhitespace=false,
+    showspaces=false,
+    showtabs=false,
+    numbers=none,
+}
 
@@
 
-\begin{frame}{Hello World}
-Hello
+\begin{frame}{Outline}
+\begin{minipage}[t][10em][t]{\linewidth}
+\tableofcontents
+\end{minipage}
 \end{frame}
 
-\begin{frame}[allowframebreaks]{Bibliography}
-\nocite{*}
+% --------------------------------------------------------------------------------------------------
+\section{Introduction}
+% --------------------------------------------------------------------------------------------------
+
+\begin{frame}[fragile]{Introduction}
+Priority Queues (PQs):
+
+\begin{itemize}
+\item Standard abstract data structure
+\item Used widely in algorithmics, operating systems, task scheduling, etc
+\item Interface consists of two $O(\log n)$ operations:
+
+\begin{lstlisting}
+void Insert(pq_t *pq, key_t k, value_t v)
+bool DeleteMin(pq_t *pq, value_t *v)
+\end{lstlisting}
+
+\item Typical backing data structures: heaps \& search trees
+\end{itemize}
+
+\end{frame}
+
+% --------------------------------------------------------------------------------------------------
+\section{Related Work}
+% --------------------------------------------------------------------------------------------------
+
+\begin{frame}{Related Work}
+\begin{itemize}
+\item Non-standard synchronization primitives
+    \begin{itemize}
+    \item \citeauthor{liu2012lock}: Array-based PQ with \lstinline|ExtractMany|
+    \item \citeauthor{israeli1993efficient}: Wait-free PQ
+    \end{itemize}
+
+\item Bounded range priorities
+    \begin{itemize}
+    \item \citeauthor{shavit1999scalable}: Combining funnels \& bins
+    \end{itemize}
+
+\item PQs in distributed memory systems
+    \begin{itemize}
+    \item \citeauthor{sanders1998randomized}: Local PQs, \lstinline|Insert| at random processor
+    \end{itemize}
+
+\item Relaxed data structures
+    \begin{itemize}
+    \item \citeauthor{kirsch2012fast}: k-FIFO queues
+    \end{itemize}
+
+\end{itemize}
+\end{frame}
+
+% --------------------------------------------------------------------------------------------------
+\section{Concepts and Definitions}
+% --------------------------------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------------------------------
+\section{Fine-grained Locking Heaps}
+% --------------------------------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------------------------------
+\section{Lock-free Priority Queues}
+% --------------------------------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------------------------------
+\section{Relaxed Priority Queues}
+% --------------------------------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------------------------------
+\section{Conclusion}
+% --------------------------------------------------------------------------------------------------
+
+% --------------------------------------------------------------------------------------------------
+\section{References}
+% --------------------------------------------------------------------------------------------------
+
+\begin{frame}[allowframebreaks]{References}
 \printbibliography

### 2014-03-24 presentation/presentation.tex
Introduction, continued

@@
 \usepackage{lmodern}
+\usepackage{textcomp}
 
@@
 
-\begin{frame}[fragile]{Introduction}
-Priority Queues (PQs):
+\begin{frame}[fragile,allowframebreaks]{Introduction}
+\acp{PQ}:
 
@@
 \item Typical backing data structures: heaps \& search trees
+\end{itemize}
+
+\framebreak
+
+\begin{itemize}
+\item In the past decade, processor clock speeds have remained the same, trend towards multiple cores
+\item New data structures required to take advantage of concurrent execution
+\item The topic of this presentation: efficient concurrent \acp{PQ}
+\item Fine-grained locking \textrightarrow ~ Lock-free \textrightarrow ~ Relaxed data structures
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[allowframebreaks]{Terminology}
+\begin{itemize}
+\item Safety conditions: nothing bad has happened yet
+    \begin{itemize}
+    \item \emph{Linearizability}: operations appear to take effect at a single point in time, the linearization point
+    \item \emph{Quiescent consistency}: in a period of quiescence, semantics equivalent to some sequential ordering
+    \item And others, i.e. \emph{sequential consistency}, \emph{serializability}
+    \end{itemize}
+
+\framebreak
+
+\item Liveness conditions: something good eventually happens
+    \begin{itemize}
+    \item \emph{Lock-freedom}: at least a single process makes progress at all times
+    \item \emph{Wait-freedom}: every process finishes in a bounded number of steps
+    \item Further liveness conditions: \emph{Starvation-freedom}, \emph{Deadlock-freedom}
+    \end{itemize}
+
+\framebreak
+
+\item \emph{Disjoint-access parallelism}: how well a data structure handles concurrent use by multiple
+      threads within disjoint areas
+\item Synchronization primitives:
+    \begin{itemize}
+    \item \ac{CAS}, \ac{FAA}, \ac{FAO}, \ac{TAS}
+    \item \ac{DCAS}, \ac{DCSS}
+    \end{itemize}
+
 \end{itemize}

### 2014-03-24 presentation/presentation.tex
Reference section title in frame headers
Fine-grained locking section

@@
 \usepackage{lmodern}
+\usepackage{nameref}
 \usepackage{textcomp}
@@
 % --------------------------------------------------------------------------------------------------
-\section{Introduction}
+\section{Introduction} \label{sec:intro}
 % --------------------------------------------------------------------------------------------------
 
-\begin{frame}[fragile,allowframebreaks]{Introduction}
+\begin{frame}[fragile,allowframebreaks]{\nameref{sec:intro}}
 \acp{PQ}:
@@
 
-\begin{frame}[allowframebreaks]{Terminology}
+\begin{frame}[allowframebreaks]{Concepts and Definitions}
 \begin{itemize}
@@
 % --------------------------------------------------------------------------------------------------
-\section{Related Work}
+\section{Related Work} \label{sec:related}
 % --------------------------------------------------------------------------------------------------
 
-\begin{frame}{Related Work}
+\begin{frame}{\nameref{sec:related}}
 \begin{itemize}
@@
 % --------------------------------------------------------------------------------------------------
-\section{Concepts and Definitions}
+\section{Fine-grained Locking Heaps} \label{sec:locking}
 % --------------------------------------------------------------------------------------------------
 
+\begin{frame}{\nameref{sec:locking}}
+\begin{itemize}
+\item Naive PQ parallelization: single global lock \textrightarrow ~ sequential bottleneck
+\item A first improvement: fine-grained locking using a lock per node
+\item \fullcite{hunt1996efficient}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:locking}}
+\framesubtitle{Innovations}
+
+\begin{itemize}
+\item One lock per node, \emph{but} additionally a global lock protecting the heap's
+      \lstinline|size| variable
+\item Insertions bottom-up, deletions top-down to reduce contention
+\item Successive insertions take disjoint paths towards the root
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:locking}}
+\framesubtitle{Limitations}
+
+\begin{itemize}
+\item A global lock remains
+\item Heap is statically allocated
+\item Frequent complex heap reorganization
+\item Disjoint-access breaks down at high traffic levels
+\item Inherent PQ bottleneck at the minimal node
+\item Benchmarks show only limited scalability up to a low thread count
+\end{itemize}
+\end{frame}
+
 % --------------------------------------------------------------------------------------------------
-\section{Fine-grained Locking Heaps}
+\section{Lock-free Priority Queues} \label{sec:lockfree}
 % --------------------------------------------------------------------------------------------------
 
+\begin{frame}{\nameref{sec:lockfree}}
+
+\end{frame}
+
 % --------------------------------------------------------------------------------------------------
-\section{Lock-free Priority Queues}
+\section{Relaxed Priority Queues} \label{sec:relaxed}
 % --------------------------------------------------------------------------------------------------
 
+\begin{frame}{\nameref{sec:relaxed}}
+
+\end{frame}
+
 % --------------------------------------------------------------------------------------------------
-\section{Relaxed Priority Queues}
+\section{Conclusion} \label{sec:conclusion}
 % --------------------------------------------------------------------------------------------------
 
+\begin{frame}{\nameref{sec:conclusion}}
+
+\end{frame}
+
 % --------------------------------------------------------------------------------------------------
-\section{Conclusion}
+\section{References} \label{sec:references}
 % --------------------------------------------------------------------------------------------------
 
-% --------------------------------------------------------------------------------------------------
-\section{References}
-% --------------------------------------------------------------------------------------------------
-
-\begin{frame}[allowframebreaks]{References}
+\begin{frame}[allowframebreaks]{\nameref{sec:references}}
 \printbibliography

### 2014-03-24 presentation/presentation.tex
Lockfree section

@@
 \begin{frame}{\nameref{sec:locking}}
+\framesubtitle{\citeauthor{hunt1996efficient}}
+
 \begin{itemize}
@@
 \begin{frame}{\nameref{sec:locking}}
-\framesubtitle{Innovations}
+\framesubtitle{\citeauthor{hunt1996efficient}: Innovations}
 
@@
 \begin{frame}{\nameref{sec:locking}}
-\framesubtitle{Limitations}
+\framesubtitle{\citeauthor{hunt1996efficient}: Limitations}
 
@@
 \begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{SkipLists}
 
+\begin{itemize}
+\item Modern concurrent PQs are mostly based on \citeauthor{pugh1990skip}'s SkipList
+\item Probabilistic ordered search structure, insertions and deletions in expected $O(\log n)$ time
+\item No reorganizations
+\item Simple implementation
+\item Excellent disjoint-access properties
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{\nameref{sec:lockfree}}
+\framesubtitle{SkipLists}
+
+\begin{itemize}
+\item Collection of linked lists with corresponding levels
+\item Lowest list contains all items, higher lists are shortcuts
+\item \lstinline|Insert| chooses a \lstinline|level| according to geometric distribution
+\end{itemize}
+
+\begin{lstlisting}
+struct slist_t {
+  size_t max_level;
+  node_t head[max_level];
+};
+struct node_t {
+  key_t key;
+  value_t value;
+  size_t level;
+  node_t *next[level];
+};
+\end{lstlisting}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{shavit2000skiplist}}
+
+\begin{itemize}
+\item First SkipList-based PQ
+\item \fullcite{shavit2000skiplist}
+\item Initially lock-based, lock-free variant published in 2008
+\item \fullcite{herlihy2012art}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{shavit2000skiplist}}
+
+\begin{itemize}
+\item Items are considered in the list once inserted on bottom level
+\item Again, insertions bottom-up and deletions top-down
+\item Deletions are split
+    \begin{itemize}
+    \item Logical deletion sets a \lstinline|deleted| flag
+    \item Physical deletion performs actual pointer manipulations
+    \end{itemize}
+\item \lstinline|DeleteMin| attempts to logically delete the head node.
+     On success: delete physically \& return node. Otherwise, continue with
+     next node.
+\item \lstinline|Insert| is equivalent to the SL insertion
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{shavit2000skiplist}}
+
+\begin{itemize}
+\item
+\end{itemize}
 \end{frame}

### 2014-03-24 presentation/presentation.tex
Remove framebreaks from concepts section
PQ, SL acronyms

@@
 
-\begin{frame}[allowframebreaks]{Concepts and Definitions}
-\begin{itemize}
-\item Safety conditions: nothing bad has happened yet
+\begin{frame}{Concepts and Definitions}
+\framesubtitle{Safety conditions: nothing bad has happened yet}
+
 \begin{itemize}
@@
 \end{itemize}
+\end{frame}
 
-\framebreak
+\begin{frame}{Concepts and Definitions}
+\framesubtitle{Liveness conditions: something good eventually happens}
 
-\item Liveness conditions: something good eventually happens
 \begin{itemize}
@@
 \end{itemize}
+\end{frame}
 
-\framebreak
+\begin{frame}{Concepts and Definitions}
+\framesubtitle{Miscellaneous}
 
+\begin{itemize}
 \item \emph{Disjoint-access parallelism}: how well a data structure handles concurrent use by multiple
@@
     \end{itemize}
-
 \end{itemize}
-
 \end{frame}
@@
     \begin{itemize}
-    \item \citeauthor{liu2012lock}: Array-based PQ with \lstinline|ExtractMany|
-    \item \citeauthor{israeli1993efficient}: Wait-free PQ
+    \item \citeauthor{liu2012lock}: Array-based \ac{PQ} with \lstinline|ExtractMany|
+    \item \citeauthor{israeli1993efficient}: Wait-free \ac{PQ}
     \end{itemize}
@@
 
-\item PQs in distributed memory systems
+\item \acp{PQ} in distributed memory systems
     \begin{itemize}
-    \item \citeauthor{sanders1998randomized}: Local PQs, \lstinline|Insert| at random processor
+    \item \citeauthor{sanders1998randomized}: Local \acp{PQ}, \lstinline|Insert| at random processor
     \end{itemize}
@@
 \begin{itemize}
-\item Naive PQ parallelization: single global lock \textrightarrow ~ sequential bottleneck
+\item Naive \ac{PQ} parallelization: single global lock \textrightarrow ~ sequential bottleneck
 \item A first improvement: fine-grained locking using a lock per node
@@
 \item Disjoint-access breaks down at high traffic levels
-\item Inherent PQ bottleneck at the minimal node
+\item Inherent \ac{PQ} bottleneck at the minimal node
 \item Benchmarks show only limited scalability up to a low thread count
@@
 \begin{itemize}
-\item Modern concurrent PQs are mostly based on \citeauthor{pugh1990skip}'s SkipList
+\item Modern concurrent \acp{PQ} are mostly based on \citeauthor{pugh1990skip}'s \ac{SL}
 \item Probabilistic ordered search structure, insertions and deletions in expected $O(\log n)$ time
@@
 \begin{itemize}
-\item First SkipList-based PQ
+\item First SkipList-based \ac{PQ}
 \item \fullcite{shavit2000skiplist}
@@
      next node.
-\item \lstinline|Insert| is equivalent to the SL insertion
+\item \lstinline|Insert| is equivalent to the \ac{SL} insertion
 \end{itemize}

### 2014-03-24 paper/paper.Rnw
Minor paper corrections

@@
 
-A crucial observation is that nodes nodes which are only partially connected
+A crucial observation is that nodes which are only partially connected
 do not affect correctness of the data structure. As soon as the first level (i.e. \lstinline|node.level[0]|)
@@
 
-Note that so far, this implementation is not linearizable: consider the case in which a
+Note that so far this implementation is not linearizable: consider the case in which a
 slow thread A is in the middle of a \lstinline|DeleteMin| call. Within this context, we refer to 

### 2014-03-24 presentation/presentation.tex
Add Shavit/Lotan figure
CAS code example
Further work on slides

@@
 \usepackage{default}
+\usepackage{graphicx}
 \usepackage[utf8]{inputenc}
@@
 
+\graphicspath{{figures/}}
 \input{../common/acronyms}
@@
 
-\begin{frame}{Concepts and Definitions}
+\begin{frame}[fragile]{Concepts and Definitions}
 \framesubtitle{Miscellaneous}
@@
 \end{itemize}
+
+\begin{lstlisting}
+bool CAS(T *ptr, T *expected, T value) {
+  if (*ptr == *expected) {
+    *ptr = value;
+    return true;
+  } else {
+    *expected = *ptr;
+    return false;
+  }
+}
+\end{lstlisting}
+
 \end{frame}
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{shavit_lotan}
+\caption{The \citeauthor{shavit2000skiplist} \ac{PQ} (Image source: \cite{shavit2000skiplist})}
+\end{figure}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{shavit2000skiplist}}
+
 \begin{itemize}
@@
 \begin{itemize}
+\item Quiescently consistent, but not linearizable
+    \begin{itemize}
+    \item Slow thread A suspended at deleted key $k$ while in \lstinline|DeleteMin|
+    \item Fast thread B first inserts $k-1$, then $k+1$
+    \item A wakes up and returns $k+1$
+    \item Linearizability would require returning $k-1$
+    \end{itemize}
+
+\item Timestamp mechanism: stamp nodes on successful insertion,
+      \lstinline|DeleteMin| ignores all stamps earlier than its own starting point
+\item Improved scalability, but heavy contention at list head
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{sundell2003fast}}
+% TODO: Way too much content - skip all except Linden?
+\begin{itemize}
+\item
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{linden2013skiplist}}
+
+\begin{itemize}
+\item
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{linden2013skiplist}}
+
+\begin{itemize}
 \item
@@
 
+% Intro relaxed
+
+\begin{itemize}
+\item
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:relaxed}}
+\framesubtitle{\citeauthor{wimmer2013data}}
+
+\begin{itemize}
+\item
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:relaxed}}
+\framesubtitle{\citeauthor{alistarhspraylist}}
+
+\begin{itemize}
+\item
+\end{itemize}
+\end{frame}
+
+% --------------------------------------------------------------------------------------------------
+\section{Results} \label{sec:Results}
+% --------------------------------------------------------------------------------------------------
+
+\begin{frame}{\nameref{sec:Results}}
+
+\begin{itemize}
+\item All benchmarks were performed on Intel Xeon processors
+\item Each thread repeatedly chooses uniformly at random between
+      \lstinline|Insert| and \lstinline|DeleteMin|
+\item Initial priority queue sizes were $2^{15}$ (Gruber), $10^6$ 
+      (\citeauthor{alistarhspraylist}), and unspecified (\citeauthor{linden2013skiplist})
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:Results}}
+\framesubtitle{Gruber}
+
+\begin{figure}
+% TODO: Sweave
+\caption{Throughput benchmark results.}
+\end{figure}
+\end{frame}
+
+\begin{frame}{\nameref{sec:Results}}
+\framesubtitle{\citeauthor{linden2013skiplist}}
+
+\begin{figure}
+\includegraphics[width=0.6 \textwidth]{results_linden}
+\caption{Throughput benchmark results.
+\emph{New}: \citeauthor{linden2013skiplist},
+\emph{HaS}: \citeauthor{shavit2000skiplist},
+\emph{SaT}: \citeauthor{sundell2003fast}. 
+(Image source: \cite{linden2013skiplist})}
+\end{figure}
+\end{frame}
+
+\begin{frame}{\nameref{sec:Results}}
+\framesubtitle{\citeauthor{alistarhspraylist}}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{results_spraylist}
+\caption{Throughput benchmark results. 
+(Image source: \cite{alistarhspraylist})}
+\end{figure}
 \end{frame}
@@
 \begin{frame}{\nameref{sec:conclusion}}
+\begin{itemize}
+\item Parallelizing \acp{PQ} is hard
+\item SkipLists currently dominate practical implementations
+\item Main limiting factor are list head accesses in \lstinline|DeleteMin|
+\item \citeauthor{linden2013skiplist} \ac{PQ} is state of the art in strict
+      semantics
+\item Much potential remains in relaxed \acp{PQ} and data structures in general
+      \textrightarrow ~ future research
+\item A standalone version of the \citeauthor{wimmer2013data} \ac{PQ} is planned
+      and will allow direct comparison against related data structures
+\end{itemize}
+\end{frame}
 
+\begin{frame}{\nameref{sec:conclusion}}
+\centering
+Questions?
 \end{frame}

### 2014-03-24 paper/paper.Rnw
Extract plotting code

--- 2014-03-20 paper/paper.tex
@@
 
+\begin{center}
+<<results = tex, echo = FALSE>>=
+\SweaveInput{../common/pqplot.Rnw}
+pqplot("../results/results.csv")
+@
+\end{center}
+
 % TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.

### 2014-03-29 presentation/presentation.Rnw
Port slides to Sweave
Various adaptions in RW, Results, etc.

--- 2014-03-24 presentation/presentation.tex
@@
 \end{lstlisting}
-
 \end{frame}
@@
 
-\item \acp{PQ} in distributed memory systems
+\item Strict \acp{PQ}
     \begin{itemize}
-    \item \citeauthor{sanders1998randomized}: Local \acp{PQ}, \lstinline|Insert| at random processor
+    \item \citeauthor{hunt1996efficient}: \textcolor{BrickRed}{Fine-grained locking heap}
+    \item \citeauthor{shavit2000skiplist}: \textcolor{BrickRed}{First SkipList-based \ac{PQ}}
+    \item \citeauthor{sundell2003fast}: \textcolor{BrickRed}{First lock-free \ac{PQ}}
+    \item \citeauthor{linden2013skiplist}: \textcolor{BrickRed}{Minimizes contention}
     \end{itemize}
@@
     \item \citeauthor{kirsch2012fast}: k-FIFO queues
+    \item \citeauthor{wimmer2013data}: \textcolor{BrickRed}{k-\ac{PQ}}
+    \item \citeauthor{alistarhspraylist}: \textcolor{BrickRed}{SprayList}
     \end{itemize}
+
+\let\thefootnote\relax\footnote{Highlighted \acp{PQ} are presented in the following}
 
@@
 % --------------------------------------------------------------------------------------------------
+
+% TODO: Figure? Whiteboard demo? 1 summary slide only?
 
@@
 \framesubtitle{\citeauthor{sundell2003fast}}
-% TODO: Way too much content - skip all except Linden?
+% TODO: Short summary only?
 \begin{itemize}
@@
 \framesubtitle{\citeauthor{linden2013skiplist}}
-
+% TODO: Short summary only?
 \begin{itemize}
@@
 \begin{frame}{\nameref{sec:relaxed}}
-\framesubtitle{\citeauthor{wimmer2013data}}
-
+\framesubtitle{\citeauthor{alistarhspraylist}}
+% Mention that large PQs are required (weak bound)
 \begin{itemize}
@@
 \begin{frame}{\nameref{sec:relaxed}}
-\framesubtitle{\citeauthor{alistarhspraylist}}
-
+\framesubtitle{\citeauthor{wimmer2013data}}
+% Essence: Local list, if necessary push to global
 \begin{itemize}
@@
 \begin{itemize}
-\item All benchmarks were performed on Intel Xeon processors
+\item Benchmarking results from selected papers \& own results
 \item Each thread repeatedly chooses uniformly at random between
       \lstinline|Insert| and \lstinline|DeleteMin|
-\item Initial priority queue sizes were $2^{15}$ (Gruber), $10^6$ 
-      (\citeauthor{alistarhspraylist}), and unspecified (\citeauthor{linden2013skiplist})
+\item We present throughput, i.e. the number of operations performed per second
 \end{itemize}
@@
 \begin{figure}
-% TODO: Sweave
-\caption{Throughput benchmark results.}
+<<results = tex, echo = FALSE>>=
+\SweaveInput{../common/pqplot.Rnw}
+pqplot("../results/results.csv")
+@
+\caption{GCC 4.8.2, 80-core Intel Xeon E7-8850 @ 2.0 GHz. \ac{PQ} initialized with $2^{15}$
+elements.}
 \end{figure}
@@
 \includegraphics[width=0.6 \textwidth]{results_linden}
-\caption{Throughput benchmark results.
+\caption{GCC 4.7.2, 32-core Intel Xeon E5-4650 @ 2.7 GHz. Initial \ac{PQ} size unknown.
 \emph{New}: \citeauthor{linden2013skiplist},
@@
 \framesubtitle{\citeauthor{alistarhspraylist}}
-
+% 10^6
 \begin{figure}
 \includegraphics[width=\textwidth]{results_spraylist}
-\caption{Throughput benchmark results. 
+\caption{GCC version unknown, 80-core Intel Xeon E7-4870 @ 2.4 GHz. \ac{PQ} initialized with $10^6$
+elements.
 (Image source: \cite{alistarhspraylist})}

### 2014-03-29 paper/paper.Rnw
Minor paper corrections

@@
 \begin{abstract}
-Priority queues are abstract data structures which store a set of (priority, item) pairs
-and allow efficient access to the item with minimal priority. Such queues are an important
+Priority queues are abstract data structures which store a set of key/value pairs
+and allow efficient access to the item with the minimal (maximal) key. Such queues are an important
 element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
@@
 
-There are several different criteria which allows reasoning about the correctness of concurrent
+There are several different criteria which allow reasoning about the correctness of concurrent
 data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
@@
 in the list simply performs an update of the associated value.
-A real-time version is also provided which will we will not discuss further (interested readers are
+A real-time version is also provided which we will not discuss further (interested readers are
 referred to \cite{sundell2003fast}).
@@
 exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
-are unused and reuse these as deletion marks. A set least significant bit signifies on a pointer signifies
+are unused and reuse these as deletion marks. A set least significant bit on a pointer signifies
 that the current node is about to be deleted.
@@
 
-This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
+This implementation uses the lock-free memory management invented by \citeauthor{valois1996lock}
 \cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
@@
 nodes \emph{after} deleted nodes. This implementation instead places the deletion flag into the
-lowest level \lstinline|next| pointer of the previous node, preventing inserations \emph{before}
+lowest level \lstinline|next| pointer of the previous node, preventing insertions \emph{before}
 logically deleted nodes.
@@
 breaking guarantees.
+
+% TODO: Mention that the global list is periodically scanned by each thread.
 
@@
       is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
-      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
-      SkipList implementation. The aim of this implementation is to minimize contention in
+      It is lock-free and uses \citeauthor{fraser2004practical}'s lock-free
+      SkipList. The aim of this implementation is to minimize contention in
       calls to \lstinline|DeleteMin|.
@@
 as they do not necessarily return the minimal element within the queue and are able to spread
-\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeable
+\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeably
 better and to larger thread counts than strict designs. Further research is necessary in order to fully

### 2014-03-29 presentation/presentation.Rnw
Sundell/Tsigas

@@
 \framesubtitle{\citeauthor{sundell2003fast}}
-% TODO: Short summary only?
+
 \begin{itemize}
-\item
+\item First lock-free \ac{PQ}, linearizable, \ac{SL}-based, distinct priorities
+\item Deletion flag packed into least significant bits of \lstinline|next| pointers
+      prevent insertion \emph{after} deleted nodes
+\item Helping mechanism ensures only a single logically deleted node exists at any time
+\item Performs significantly better than the \citeauthor{hunt1996efficient} heap, slightly better
+      than a SkipList protected by a global lock
+\item \fullcite{sundell2003fast}
 \end{itemize}

### 2014-03-29 presentation/presentation.Rnw
Linden slides

@@
 \framesubtitle{\citeauthor{linden2013skiplist}}
-% TODO: Short summary only?
+
 \begin{itemize}
-\item
+\item Most efficient strict \ac{PQ}, \ac{SL}-based \& linearizable
+\item Concurrent strict \ac{PQ} performance limited by contention and \ac{CAS} failures in
+      \lstinline|DeleteMin| \textrightarrow ~ Minimize \ac{CAS} calls
+\item Deleted nodes form prefix of \ac{SL}, deletion flag packed into \lstinline|next| pointer of
+      \emph{previous} node prevents insertion \emph{before} deleted node
+\item Most \lstinline|DeleteMin| perform logical deletion (1 \ac{CAS}) only, physical deletion
+      only once a bound is reached
+\item Improves upon best previous \acp{PQ} by up to factor 2
+\item \fullcite{linden2013skiplist}
 \end{itemize}
@@
 
-\begin{itemize}
-\item
-\end{itemize}
+\begin{figure}
+\includegraphics[width=\textwidth]{linden}
+\caption{The \citeauthor{linden2013skiplist} \ac{PQ}. Concurrent \lstinline|Insert(3)| and
+\lstinline|DeleteMin| operations. Top right: \lstinline|DeleteMin| succeeds first,
+\lstinline|Insert(3)| \ac{CAS} fails. Bottom right: \lstinline|Insert(3)| succeeds first,
+\lstinline|DeleteMin| returns 3. (Image source: \cite{linden2013skiplist})}
+\end{figure}
 \end{frame}

### 2014-03-29 presentation/presentation.Rnw
Relaxed slides

@@
 \item \fullcite{shavit2000skiplist}
-\item Initially lock-based, lock-free variant published in 2008
+\item Initially lock-based (linearizable), lock-free (quiescently consistent)
+      variant published in 2008
 \item \fullcite{herlihy2012art}
@@
 
-% Intro relaxed
-
 \begin{itemize}
-\item
+\item Strict \acp{PQ} have inherent bottleneck at minimal element
+\item To improve further: $< 1$ \ac{CAS} per \lstinline|DeleteMin|
+\item Another approach is to relax semantics, i.e. instead of returning \emph{the} minimal element,
+      return one of the $k$ minimal elements
 \end{itemize}
@@
 \framesubtitle{\citeauthor{alistarhspraylist}}
-% Mention that large PQs are required (weak bound)
+
 \begin{itemize}
-\item
+\item Relaxed \ac{SL}-based \ac{PQ}, safety properties unclear
+\item \lstinline|DeleteMin| returns one of the $O(P \log^3 P)$ elements
+\item Degrades to random-remove if the \ac{PQ} is small compared to thread count $P$ (for $P = 80$,
+      $P \log^3 P \approx 7000$)
+\item \lstinline|DeleteMin| performs random walk: starting at the list head on some level $l$,
+      repeatedly follow a randomized number of \lstinline|next[l]| pointers, descend a randomized
+      number of levels
+\item Parameters are chosen s.t. each element within the walk has approximately equal probability
+      of being returned
+\item Benchmarks show scalability comparable to a random-remove \lstinline|Delete| up to at
+      least 80 threads
+\item \fullcite{alistarhspraylist}
 \end{itemize}
@@
 \begin{itemize}
-\item
+\item First relaxed linearizable \acp{PQ}: we discuss the hybrid k-\ac{PQ} which, provides a bound of
+      $kP$ missed elements in \lstinline|DeleteMin|
+\item Consists of a list of globally visible elements, and per thread: a local item list, and a local
+      \ac{PQ}
+\item \lstinline|Insert| accesses only the local structures as long as guarantees are not violated,
+      otherwise the global list is updated the the local view is synchronized
+\item \lstinline|DeleteMin| pops the local queue if it is non-empty; otherwise spy, i.e. copy
+      elements from another thread's local list
+\item Very good scalability up to 10 threads, further limited gains with rising thread counts
+\item \fullcite{wimmer2013data}
 \end{itemize}
+\end{frame}
+
+\begin{frame}{\nameref{sec:lockfree}}
+\framesubtitle{\citeauthor{wimmer2013data}}
+
+\begin{figure}
+\includegraphics[width = 0.65\textwidth]{wimmer}
+\caption{The \citeauthor{wimmer2013data} hybrid k-\ac{PQ}. (Image source: \cite{wimmer2013data})}
+\end{figure}
 \end{frame}

### 2014-03-29 src/noble.cpp
Allocate noble keys/values in advance

@@ -1,9 +1,18 @@
 #include <noble.h>
 
+static constexpr uint32_t K_STRIDE = 1 << 12;
+static constexpr uint32_t K_SIZE   = 1 << 20;
+
 Noble::Noble()
 {
     /* Lock-free, bounded memory usage. */
     m_q = pq_t::CreateLF_EB();
+
+
+    m_keys = new uint32_t[K_SIZE];
+    for (uint32_t i = 0; i < K_SIZE; i++) {
+        m_keys[i] = i * K_STRIDE;
+    }
 }
 
 Noble::~Noble()
@@ -15,13 +24,13 @@ Noble::~Noble()
     } while (nonempty);
 
     delete m_q;
+    delete[] m_keys;
 }
 
 void
 Noble::insert(const uint32_t v)
 {
-    uint32_t *u = new uint32_t { v };
-    m_q->Insert(u, u);
+    m_q->Insert(&m_keys[v / K_STRIDE], &m_keys[v / K_STRIDE]);
 }
 
 bool
@@ -39,7 +48,6 @@ Noble::delete_min(uint32_t &v)
     }
 
     v = *u;
-    delete u;
 
     return true;
 }

### 2014-03-29 src/noble.h

@@ -18,5 +18,6 @@ private:
     typedef NBL::PQueue<uint32_t, uint32_t> pq_t;
 
     pq_t *m_q;
+    uint32_t *m_keys;
 };
 #endif /* __NOBLE_H */

### 2014-03-29 paper/paper.Rnw
Benchmark notes
Results section

@@
 \begin{itemize}
-\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
+\item \textit{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
       by the C++ standard library, protected by a single global lock.
-\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
+\item \textit{Heap} \citeauthor{hunt1996efficient} provide an implementation
       of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
@@
       instead.
-\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
+\item \textit{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
       is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
-      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
+      \textit{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
       According to the authors, the commercial Noble library includes a more efficient
       version of this data structure that can also handle duplicate priority values.
-\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
+\item \textit{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
       is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
@@
       calls to \lstinline|DeleteMin|.
-\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
+\item \textit{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
       SkipList using random walks to spread data accesses
@@
 
-All benchmarks were compiled with \verb|-O3| using GCC 4.8.2. Evaluations took place
-on an 80-core Intel Xeon machine. The benchmarking code was adapted from \citeauthor{linden2013skiplist}'s
+All benchmarks were compiled with \verb|-O3| using GCC 4.8.2, and executed with threads pinned to
+cores. Evaluations took place
+on an 80-core Intel Xeon E7-8850 machine with each processor clocked at 2 GHz.
+The benchmarking code was adapted from \citeauthor{linden2013skiplist}'s
 benchmarking suite and is available at \url{https://github.com/schuay/seminar_in_algorithms}.
@@
 
+The \textit{Linden} priority queue emerged as the clear winner of these benchmarks,
+with impressive scalability while all threads remain on the same socket. In fact, it is the only
+tested implementation to scale to any significant degree; \textit{Heap} achieves minor gains
+for up to 3 threads, while all others immediately lose throughput under concurrency.
+Performance of the \textit{Linden} queue drops significantly once it is executed on more than 10
+threads, thus incurring communication overhead across sockets. This effect is repeated to a lesser
+degree every time an additional socket becomes active. However, the \textit{Linden} queue has
+clearly superior throughput until the tested maximum of 80 threads.
+
+\textit{GlobalLock} also performs surprisingly well at high concurrency levels.
+It remains competitive when compared to \textit{Linden} at about $\frac{2}{3}$ of its throughput.
+For single-threaded execution, it achieves the highest throughput of all tested implementations.
+It is interesting to note that the \textit{GlobalLock} shows complementary behavior to
+\textit{Linden} when new sockets become active; we believe that the additional latency incurred
+by cross-socket communications might lead to reduced contention at the single lock.
+
+All remaining implementations behave similarly, displaying very low throughput under concurrency.
+This is a surprise in the case of \textit{Noble}; the original benchmarks performed by
+\citeauthor{sundell2003fast} led us to expect improvements upon both\textit{GlobalLock}
+and \textit{Heap} \cite{sundell2003fast}. However, we are somewhat reassured in our benchmarks by
+the fact that \citeauthor{linden2013skiplist}'s results are similar.
+
+The \textit{SprayList} result is even more unexpected, and we believe that it may be caused by
+our use of \lstinline|malloc| instead of the included custom lock-free allocator, which we were
+unable to execute without errors.
+
+As a final observation we note that even the most efficient strict concurrent priority queue
+never exceeds the throughput of the simple sequential heap executed on a single thread by more
+than a factor of two. There is hope however since relaxed data structures continue to scale
+beyond 10 threads (see the benchmarks performed in \cite{alistarhspraylist} for an example).
+
 % TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
@@
 % Drop-off when using more than 1 socket, Linden p17.
+% Linden drop-off for new sockets, global-lock improvement
 

### 2014-03-30 src/pqbench.cpp
Increase default heap size
This helps avoid quasi-deadlocks at high concurrency.

@@ -23,7 +23,7 @@ static std::atomic<bool> loop;
 static std::atomic<int> wait_barrier;
 
 static GlobalLock pq_globallock;
-static Heap pq_heap(DEFAULT_SIZE << 2);
+static Heap pq_heap(DEFAULT_SIZE << 3);
 static Noble pq_noble;
 static Linden pq_linden(DEFAULT_OFFSET);
 static SprayList pq_spraylist;

### 2014-03-30 src/pqbench.cpp
Set the default linden BoundOffset to 32

@@ -15,7 +15,7 @@
 
 #define DEFAULT_SECS     (10)
 #define DEFAULT_NTHREADS (1)
-#define DEFAULT_OFFSET   (128)
+#define DEFAULT_OFFSET   (32)
 #define DEFAULT_SIZE     (1 << 15)
 #define DEFAULT_VERBOSE  (false)
 

### 2014-03-30 paper/paper.Rnw
Parameter details for heap/linden benchmarks
Update BoundOffset in paper

@@
       use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
-      instead.
+      instead. The heap capacity was $2^{18}$ --- lower values led to a quasi-deadlock situation under high concurrency
+      as described in \cite{dragicevic2008survey}.
 \item \textit{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
@@
       SkipList. The aim of this implementation is to minimize contention in
-      calls to \lstinline|DeleteMin|.
+      calls to \lstinline|DeleteMin|. We chose $32$ as the \lstinline|BoundOffset| in order to optimize
+      performance on a single socket. A \lstinline|BoundOffset| of $128$ performed only marginally better
+      at high thread counts.
 \item \textit{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s

### 2014-03-31 src/spraylist.cpp
Use TRANSACTIONAL argument in spraylist insert
Same as the original spraylist benchmark.

@@ -51,7 +51,7 @@ SprayList::init_thread(const size_t nthreads)
 void
 SprayList::insert(const uint32_t v)
 {
-    sl_add(m_q, v, READ_ADD_REM_ELASTIC_TX);
+    sl_add(m_q, v, TRANSACTIONAL);
 }
 
 bool

### 2014-03-31 paper/paper.Rnw
Remove comments from paper

@@
 \maketitle
-
-\begin{comment}
-Abstract formula
-----------------
-
-1) big picture problem or topic widely debated in your field.
-2) gap in the literature on this topic.
-3) your project filling the gap.
-4) the specific material that you examine in the paper.
-5) your original argument.
-6) a strong concluding sentence.
-
-Introduction formula
---------------------
-
-A good paper introduction is fairly formulaic. If you follow a simple set of
-rules, you can write a very good introduction. The following outline can be
-varied. For example, you can use two paragraphs instead of one, or you can
-place more emphasis on one aspect of the intro than another. But in all cases,
-all of the points below need to be covered in an introduction, and in most
-papers, you don't need to cover anything more in an introduction.
-
-Paragraph 1: Motivation. At a high level, what is the problem area you are
-working in and why is it important? It is important to set the larger context
-here. Why is the problem of interest and importance to the larger community?
-
-Paragraph 2: What is the specific problem considered in this paper? This
-paragraph narrows down the topic area of the paper. In the first paragraph you
-have established general context and importance. Here you establish specific
-context and background.
-
-Paragraph 3: "In this paper, we show that ...". This is the key paragraph in
-the intro - you summarize, in one paragraph, what are the main contributions of
-your paper given the context you have established in paragraphs 1 and 2. What
-is the general approach taken? Why are the specific results significant? This
-paragraph must be really really good. If you can't "sell" your work at a high
-level in a paragraph in the intro, then you are in trouble. As a reader or
-reviewer, this is the paragraph that I always look for, and read very
-carefully.
-
-You should think about how to structure this one or two paragraph summary of
-what your paper is all about. If there are two or three main results, then you
-might consider itemizing them with bullets or in test (e.g., "First, ..."). If
-the results fall broadly into two categories, you can bring out that
-distinction here. For example, "Our results are both theoretical and applied in
-nature. (two sentences follow, one each on theory and application)"
-
-Paragraph 4: At a high level what are the differences in what you are doing,
-and what others have done? Keep this at a high level, you can refer to a future
-section where specific details and differences will be given. But it is
-important for the reader to know at a high level, what is new about this work
-compared to other work in the area.
-
-Paragraph 5: "The remainder of this paper is structured as follows..." Give the
-reader a roadmap for the rest of the paper. Avoid redundant phrasing, "In
-Section 2, In section 3, ... In Section 4, ... " etc.
-
-A few general tips:
-
-Don't spend a lot of time into the introduction telling the reader about what
-you don't do in the paper. Be clear about what you do do, but don't dwell here
-on what you don't do.  Does each paragraph have a theme sentence that sets the
-stage for the entire paragraph? Are the sentences and topics in the paragraph
-all related to each other?  Do all of your tenses match up in a paragraph?
-\end{comment}
 
@@
 
-% TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.
-% Better hardware description - see SprayList section 4, Linden p16.
-% Note: Strong degradation of noble PQ with growing initial size.
-% Standard PQ benchmarking methodology, cite linden + ??.
-% Drop-off when using more than 1 socket, Linden p17.
-% Linden drop-off for new sockets, global-lock improvement
-
-\begin{comment}
-Sections / rough structure:
-* Basic concepts and definitions. Linearizability, sequ./quiescent consistency,
-  lock-free, wait-free, disjoint-access parallelism (one of the papers has good
-  summaries of these).
-  Maybe atomic primitives such as CAS (but probably not).
-* Priority queue definitions, semantics, usages. Mention inherent
-  non-scalability through DeleteMin().
-* Skiplist, heap definitions, semantics, usages.
-* Follow development of current state of the art from Hunt Heap ~> Shavit ~>
-  Tsigas ~> Linden.
-* Maybe benchmarks on mars.
-
-Additional notes:
-* Sketch development: Heap, Fraser Skiplist (Details, mention why linearizable),
-  Sundell/Tsigas, other Skiplist-based PQs
-* Modern developments: SprayList, Wimmer et al. Relation to other relaxed structures such as k-FIFO.
-\end{comment}
-
 \section{Related Work} \label{sec:related}
@@
 
-\begin{comment}
-  * hunt: most efficient old-school algorithm
-  * lots of other older references in [4]
-  * israeli, rappoport: wait-free, non-available atomic primitive
-  * lotan, shavit [4]: skiplist, lock-based. first to propose usage of skip-lists [11]
-  * sundell, tsigas [3]: skiplist, lock-free, linearizable, unique priorities
-  * herlihy, shavit in art of multiprocessor programming: based on [4], lock-free.
-  * linden, jonsson [11]: skiplist, lock-free, linearizable, reduced mem contention.
-    first lock-free PQ algorithm [11, 12]
-  * specialized versions such as
-    * bounded priorities: [7] and others
-    * probabilistic extractMin, extractMany: [9]
-\end{comment}
-
 \section{Conclusion} \label{sec:conclusion}
@@
 
-\begin{comment}
-http://leo.stcloudstate.edu/acadwrite/conclude.html
-http://writingcenter.unc.edu/handouts/conclusions/
-http://www.wikihow.com/Write-a-Conclusion-for-a-Research-Paper
-\end{comment}
-
 \printbibliography

### 2014-03-31 paper/paper.Rnw
Explain why S+L, Wimmer PQs were not benched

@@
 
+Unfortunately, we were not able to obtain an implementation of the \citeauthor{shavit2000skiplist}
+priority queue for benchmarking. The \citeauthor{wimmer2013data} data structure was omitted in the
+following benchmarks since it is tightly coupled with the task-scheduling framework \emph{Pheet},
+and cannot directly be compared to other implementations in its current form.
+
 In each test run, the examined priority queue was initially filled with $2^{15}$

### 2014-04-04 paper/paper.Rnw
Paper tweaks

@@
 
-\title{State of the Art in Concurrent Priority Queues \\
+\title{Practical Concurrent Priority Queues \\
        Student Paper for Seminar in Algorithms 2013W \\
@@
 % Priority queues.
-
+% TODO: Check for whitespace error @ beginning of line this paragraph.
 Priority queues are an abstract data structure allowing insertion of items with a given priority
 and removal of the lowest-priority item in logarithmic time. Search trees and Heaps (which are
-flattened representations of complete trees such that each node's value is at least as large as
+flattened representations of complete trees such that each node's key is at least as large as
 those of both children) are concrete data structures which are usually used to implement sequential
@@
 bool slist_delete(slist_t *l, key_t k, value_t &v);
+bool slist_contains(slist_t *l, key_t k);
 \end{lstlisting}
@@
 every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
-Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heap
+Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heaps
 is fixed upon creation.
@@
 In contrast to other structures covered in this paper, this priority queue is restricted to
-contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
+contain items with distinct priorities. Inserting a new item with a priority already contained
 in the list simply performs an update of the associated value.
@@
 This is a surprise in the case of \textit{Noble}; the original benchmarks performed by
-\citeauthor{sundell2003fast} led us to expect improvements upon both\textit{GlobalLock}
+\citeauthor{sundell2003fast} led us to expect improvements upon both \textit{GlobalLock}
 and \textit{Heap} \cite{sundell2003fast}. However, we are somewhat reassured in our benchmarks by

### 2014-04-04 paper/paper.Rnw
Clarify that quiescent consistency is composable as well

@@
 While early efforts have focused mostly on parallelizing Heap structures
-\cite{hunt1996efficient}, % TODO: Many more, see \cite{shavit2000skiplist}
+\cite{hunt1996efficient},
 more recently priority queues based on \citeauthor{pugh1990skip}'s SkipLists
@@
 data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
-effect at a single instant in time at so-called linearization points. Linearizability is composable ---
-any data structure composed of linearizable objects is also linearizable.
+effect at a single instant in time at so-called linearization points.
+\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
+guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
+after a period of quiescence, i.e. an interval without active operations has passed.
+Linearizability as well as quiescent consistency are composable ---
+any data structure composed of linearizable (quiescently consistent) objects is also linearizable
+(quiescently consistent).
 \emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
@@
 operations.
-\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
-guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
-after a period of quiescence, i.e. an interval without active operations has passed.
 

### 2014-04-04 paper/paper.Rnw
Add section on synchronization primitives

@@
 
+% Synchronization primitives.
+
+Lock-free algorithms and data structures are commonly constructed using synchronization primitives
+such as \acf{CAS}, \ac{FAA}, \ac{FAO}, and \ac{TAS}. The \ac{CAS} instruction, which atomically
+compares a memory location to an expected value and sets it to a new value if they are equal,
+is implemented on most modern architectures and can be considered a basic building block of lock-free
+programming. More exotic primitives such as \ac{DCAS} and \ac{DCSS} exist as well, but are not yet
+widely available and require inefficient software emulations to be used.
+
 % Disjoint-access parallelism.
@@
 Contention is a limiting factor regarding scalability: concurrent reads and writes to the same
-location must be serialized by the cache coherence protocol, and only a single concurrent \ac{CAS} % TODO: Short section on primitives.
+location must be serialized by the cache coherence protocol, and only a single concurrent \ac{CAS}
 can succeed while all others must retry. \emph{Disjoint-access parallelism} is the concept of

### 2014-04-04 paper/paper.Rnw
Fix whitespace issue after \lstinline

@@
 % Priority queues.
-% TODO: Check for whitespace error @ beginning of line this paragraph.
+
 Priority queues are an abstract data structure allowing insertion of items with a given priority
@@
 those of both children) are concrete data structures which are usually used to implement sequential
-priority queues. However, both require fairly elaborate reorganization after \lstinline|DeleteMin|
+priority queues. However, both require fairly elaborate reorganization after
+{\lstset{breakatwhitespace=true} \lstinline|DeleteMin|}
 and/or \lstinline|Insert| operations, which are especially challenging to achieve in a concurrent

### 2014-04-04 paper/paper.Rnw
Remove TODOs that are really WONTDOs

@@
 they are both simple to implement, and exhibit excellent disjoint-access parallelism. In contrast to
-search trees and Heaps, reorganization is not necessary, since SkipLists rely on
+search trees and Heaps, reorganization is not necessary since SkipLists rely on
 randomization for an expected $O(\log n)$ complexity of \lstinline|Insert|, \lstinline|Search|
@@
 according to a geometric distribution; deletion simply removes the node.
-
-% TODO: Maybe integrate into description - otherwise remove. Do we need further detail?
 
@@
 
-% TODO: This counts for all skiplist impls.
-
 A crucial observation is that nodes which are only partially connected
@@
 simply ignores all nodes it sees that have not been fully inserted at the time it was called.
-
-% TODO: Linearizability points.
 
@@
 
-% TODO: Mention backoff.
-
 This implementation uses the lock-free memory management invented by \citeauthor{valois1996lock}
@@
 linearization points for all possible outcomes of all operations.
-
-% TODO: Further details?
 
@@
 operations within this context implicitly choose a key uniformly at random from
-the range of all 32-bit integers. This methodology seems to be the de facto standard for concurrent % TODO: Similar phrasing as linden :(
+the range of all 32-bit integers. This methodology seems to be the de facto standard for concurrent
 priority queue benchmarks

### 2014-04-04 paper/paper.Rnw
Mention insertion flag, memory for Linden queue

@@
 
-% TODO: Mention inserting flag and memory management!
-
-In this implementation, \lstinline|DeleteMin| operations only perform logical deletion by setting
+In this implementation, most \lstinline|DeleteMin| operations only perform logical deletion by setting
 the deletion flag with a single \ac{FAO} call; nodes are only deleted physically once a certain
@@
 Insertions occur bottom-up and follow the basic \citeauthor{fraser2004practical} algorithm \cite{fraser2004practical},
-taking the separation of deletion flags and nodes into account.
+taking the separation of deletion flags and nodes into account. The \lstinline|node.inserting| flag
+is set until the node has been fully inserted, and prevents moving the list head past a partially
+inserted node. \citeauthor{fraser2004practical}'s epoch-based reclamation scheme \cite{fraser2004practical}
+is used for memory management.
 
